Plant Arbovirus Mobilizes a Vector Salivary Protein to Initiate Plant Infection

Version 1

Abstract

Abstract

Plant arboviruses rely heavily on insects’ feeding activities for successful transmission. Insect salivary proteins have been suggested to be essential for successful viral infection, but their exact mechanisms are largely unknown. In this study, we reveal that salivary factors from Laodelphax striatellus are necessary for infection of Rice stripe virus (RSV) in plants. A salivary carbonic anhydrase (LssaCA) is identified as an essential factor in promoting RSV infection. LssaCA interacts with a rice thaumatin-like protein (OsTLP) that has endo-β-1,3-glucanase activity and can degrade callose in plants. RSV infection induces callose deposition, which can be reversed by LssaCA. Furthermore, LssaCA directly binds to the RSV nucleocapsid protein (NP) in salivary glands, and the LssaCA-RSV NP complex still binds OsTLP and further increases its glucanase activity. This study provides new insights into the tripartite virus-insect vector-plant interaction, which is relevant to many agriculturally important plant arboviruses whose transmission is facilitated by insect salivary proteins.

Introduction

During plant virus transmission by piercing-sucking insects, arboviruses are inoculated into the plant phloem via the secreted insect’s saliva (Arca & Ribeiro, 2018; Conway et al., 2016; Wu, Yi, Lv, Mao, & Wei, 2022). Thus, insect saliva acts as an interface for the virus-insect-host tripartite interaction and it can directly promote pathogen transmission to, and infection of, the hosts (Sun et al., 2020; Wu et al., 2022). However, despite the importance of insect salivary proteins in the tripartite interaction, there is still much to learn about how these proteins enable successful viral infection.

Previous studies have revealed that there are two ways in which insect saliva facilitates viral infection (Arca & Ribeiro, 2018). One is an indirect approach whereby the saliva modulates the host microenvironment at the feeding site, and saliva effectors work together to allow the arthropod to go unnoticed while it feeds on the host plant (Acevedo et al., 2019; Arca & Ribeiro, 2018; Sun et al., 2020). The other is direct regulation, whereby saliva proteins promote virus transmission through specific molecular interactions (Wen et al., 2019). Our recent work indicated that a Laodelphax striatellus mucin protein, LssaMP, enables the formation of the salivary sheath and facilitates the transmission of rice stripe virus (RSV) into the rice phloem (Huo et al., 2022). A recent study on leafhoppers revealed that the expression of a saliva calcium-binding protein is inhibited by rice gall dwarf virus (RGDV), thus causing an increase of cytosolic Ca2+ levels in rice and triggering callose deposition and H2O2 production. This increases the frequency of insect probing, thereby enhancing viral horizontal transmission into the rice phloem (Wu et al., 2022). Direct saliva protein-pathogen interactions have been reported in animal pathogens. For example, during transmission of Borrelia burgdorferi by Ixodes scapularis, the saliva protein Salp15 of I. scapularis binds to the bacterial outer surface protein C, which prevents the bacterium from being recognized by the animal immune system. In this way, the saliva protein enables the pathogen to infect the animal host (Ramamoorthi et al., 2005; Schuijt et al., 2008). The Anopheles gambiae saliva sporozoite-associated protein (AgSAP) binds to the surface of the Plasmodium berghei sporozoite. The binding of AgSAP to heparan sulfate inhibits local inflammatory responses, thereby facilitating early Plasmodium infection in the vertebrate host. Although most plant viruses are completely dependent on insect vectors for plant-to-plant transmission (Gray, 2008), the direct function of insect saliva proteins in mediating virus transmission remains largely uninvestigated.

During sap-feeding, arthropods produce two distinct types of saliva at different stages of the feeding process; gel saliva and watery saliva (Bonaventure, 2012; Lou et al., 2019). The former forms a salivary sheath to provide a smooth path for the stylet penetration (Lou et al., 2019). The latter is secreted into the phloem sieve elements to prevent them from plugging up, suppresses plant defense responses, induces pattern-triggered immunity (PTI) through herbivore-associated molecular patterns (HAMPs), and induces effector-triggered immunity (ETI) through avirulent gene, etc. (Huang, Liu, Xu, Bao, & Zhang, 2017; Ji et al., 2017; Yi, Wu, & Wei, 2021).

For phloem-feeding insects, callose deposited on phloem sieve tubes can function as a defense mechanism by reducing insect feeding and preventing long-distance viral movement (Hao et al., 2008; Hipper, Brault, Ziegler-Graff, & Revers, 2013; Will & Vilcinskas, 2015; Zavaliev, Ueki, Epel, & Citovsky, 2011). Callose is a β-(1,3)-D-glucan polysaccharide that is synthesized by callose synthases and degraded by β-(1,3)-glucanases. Plants can defend themselves by depositing callose at the sieve tubes in response to virus infection, whereas viruses can counter this by activating β-(1,3)-glucanases to degrade callose and unplug the sieve tube occlusions (Bucher et al., 2001; Hao et al., 2008; Wu et al., 2022; Zavaliev et al., 2011).

RSV is the causative agent of rice stripe disease, a serious disease of rice crops that has occurred repeatedly in China, Japan, and Korea (Xu, Fu, Tao, & Zhou, 2021). RSV is completely dependent on insect vectors for transmission to its host plants, and L. striatellus is the main vector (Xu et al., 2021; Zhao et al., 2018). L. striatellus transmits RSV in a persistent-propagative manner. The virus initially infects the midgut, then disperses from the hemolymph into the salivary glands, and is inoculated into the plant host during L. striatellus feeding (Huo et al., 2022). L. striatellus belongs to the order Hemiptera, whose members mainly feed from sieve tubes through their mouthparts (stylets) that penetrate plant tissues and reach sieve tubes to ingest the phloem sap (Tjallingii, 2006; van Bel & Will, 2016). RSV is secreted into the rice phloem via the watery saliva (Huo et al., 2022; Wang & Blanc, 2021).

In this study, we identified a molecular interaction among RSV, an L. striatellus saliva protein, and a plant β-1,3-glucanase. The insect saliva protein directly binds to the RSV nucleocapsid protein (NP) and then binds to a rice thaumatin-like protein to activate its β-1,3-glucanase activity. The activation of β-1,3-glucanase helps facilitate RSV infection of the host plant by inhibiting callose deposition in response to viral infection.

Results

L. striatellus salivary carbonic anhydrase is an essential factor in enhancing RSV infection

To investigate the influence of L. striatellus salivary proteins on RSV infection levels in rice plants, two experiments were conducted and compared: viruliferous insect feeding (VIF) and microinjection of purified virus (MPV) (Fig. 1A). The results showed that VIF caused high RSV infection levels and obvious rice stripe disease symptoms, while MPV did not lead to disease symptoms, and only a low level of RSV NP protein was detected inside the plant (Fig. 1B-E). These results suggested that the insect saliva factors are necessary for enabling RSV infection.

Laodelphax striatellus feeding is necessary for Rice stripe virus (RSV)to infect rice.

A. The schematic diagram shows the two methods of rice inoculation with RSV: viruliferous insect feeding (VIF) and microinjection of purified virus (MPV). Three-leaf rice seedlings were used and viral infection was detected three weeks after inoculation.

B-D. Plant symptoms of rice stripe disease were observed in the VIF group (B-D), including heart leaf curling (C) and the striped spots on the leaves (D), but not in the MPV group (B and D). C shows an enlarged image of the VIF-group leaves.

A. E. Western blots to show the levels of RSV infection in plants inoculated with VIF or MPV. Anti-RSV antibodies were used to detect the RSV titers in the plants.

To identify the salivary proteins that play roles in mediating RSV infection, we have sequenced the L. striatellus saliva proteins and focused on the watery proteins that were inoculated into the plant phloem sieve tubes together with RSV (Huo et al., 2022). A salivary protein, Sap6 (GenBank accession No. RZF48846.1), was identified to have a high expression level in L. striatellus salivary glands (Fig. 2A). Western blotting assay showed that Sap6 was secreted into the plant as a watery saliva protein (Fig. 2B-C). Sequence homology analysis revealed Sap6 as a carbonic anhydrase-like protein, and enzyme active site analysis and enzyme activity experiment indicated Sap6 as a carbonic anhydrase (Fig. 2D-E). Sap6 was then named L. striatellus salivary carbonic anhydrase (LssaCA).

Characteristics of the L. striatellus saliva carbonic anhydrase (LssaCA).

A. RT-qPCR analyses of LssaCA transcript levels in different tissues. R-body, the remaining body after removal of the guts, salivary glands, fat bodies, and ovaries.

B. Western blots showing LssaCA protein distribution in L. striatellus salivary glands (SBPH SG) and rice plants (Rice). The weaker signal in dsLssaCA-treated SBPH indicates a specific knockdown of LssaCA by dsLssaCA. Rice plants fed by SBPH acquire LssaCA (+SBPH), whereas control plants do not contain LssaCA protein (-SBPH).

C. Western blots showing the distribution of LssaCA in watery and gel saliva. Total watery or gel saliva proteins were detected by silver staining.

D. Schematic diagram of LssaCA protein sequence. SP, signal peptide; Carb_anhydrase domain, conserved eukaryotic-type carbonic anhydrase sequence. Triangles indicate seven predicted catalytically active residues.

E. Esterase activity of recombinantly expressed LssaCA protein.

To determine the role of LssaCA in RSV infection of rice plants, a LssaCA-specific double-stranded RNA (dsLssaCA) was synthesized and delivered into the hemocoel of RSV-infected 3rd instar nymphs to interfere with the gene expression. At 2 days post-microinjection (dpi), the LssaCA mRNA levels were measured by RT-qPCR and compared with the dsGFP control. The results showed that dsLssaCA treatment reduced the LssaCA mRNA level by 80% (Fig. 3A). The two groups of insects were then allowed to feed on healthy rice seedlings (5 insects per rice seedling) for 2 days. The RSV NP RNA levels (indicative of RSV titers) in the plants were measured at 14 days post-feeding (dpf). Compared with the rice seedlings fed by dsGFP-treated insects, those fed by dsLssaCA-treated insects had significantly reduced RSV titers (Fig. 3B), indicating that LssaCA played an essential role in mediating RSV infection of the rice plants.

The influence of LssaCA deficiency on RSV transmission and viral infection of plants.

A. The efficiency of gene silencing as determined by RT-qPCR. Each dot represents salivary glands from five RSV-infected third-instar nymphs.

B. RSV infection level in rice plants as determined by RT-qPCR. RSV infection levels are represented by the copy number of the NP gene. Each dot represents one rice plant fed on for 2 d and then grown for another 14 d.

C. RSV titers in L. striatellus salivary glands as determined by RT-qPCR. Each dot represents salivary glands from five RSV-infected third-instar nymphs.

D. RSV titers in L. striatellus saliva as determined by RT-qPCR. Each dot represents one saliva sample from 10 insects fed with an artificial diet.

E. RSV titers in rice plants as determined by RT-qPCR. Rice plants were fed for 24 h and RSV titers were assayed immediately post-feeding. Each dot represents one rice plant.

F. Schematic diagram to show the process of microinjecting an RSV particle solution into a plant. The RSV was pre-incubated with BSA or recombinantly expressed LssaCA protein before being microinjected into the plant phloem.

G. RT-qPCR to determine the level of RSV infection in microinjection-inoculated rice. The titers of RSV (copy number of NP) in rice plants were measured at 14 days post-inoculation (dpi). BSA was used as a negative control. *, p < 0.05.

DsLssaCA and dsGFP indicate LssaCA- and GFP-specific dsRNA, respectively. ****, p < 0.0001; **, p < 0.01; ns, not significant.

LssaCA enhances RSV infection not before the viruses have been inoculated into the plant phloem

To determine at which step LssaCA-deficiency affected RSV infection, we analyzed the influence of LssaCA-deficiency on RSV titers during several steps of the RSV transmission process, including the virus load in the salivary glands and saliva, and the initial inoculation levels in planta. First, at 2 days after dsLssaCA treatment of the viruliferous insects when LssaCA was significantly downregulated, RT-qPCR was used to detect RSV titer in the insect salivary glands and the results showed that LssaCA deficiency did not reduce RSV titers in this tissue (Fig. 3C). Then, the LssaCA-deficient insects were allowed to feed on an artificial diet and their saliva was collected for RSV level analysis, which revealed that LssaCA-deficiency did not reduce the RSV titer in the secreted saliva (Fig. 3D). Finally, plants were fed with 10 viruliferous insects for 24 h respectively and the RSV titers were measured immediately after feeding, which showed that LssaCA-deficiency did not affect the RSV initial inoculation levels in planta (Fig. 3E). These results suggested that LssaCA promoted RSV infection by a mechanism occurred not in insects or early stage of viral entry in plants, but in planta after viral inoculation.

As shown in Fig. 1, VIF but not MPV is able to enable RSV infection of rice plants, suggesting the necessity of insect saliva factors in enabling RSV infection. To further confirm whether LssaCA played a role in directly enhancing RSV levels in planta, we produced a recombinant LssaCA protein to pre-incubate with purified RSV particle solution and used bovine serum albumin (BSA) as a control for comparison. The two groups were then microinjected into leaves on different plants, and at 14 dpi their viral titers were measured using NP gene-specific RT-qPCR. The results showed that compared to those inoculated with BSA plus RSVs particles, those inoculated with LssaCA plus these same particles exhibited significantly higher RSV titer values (Fig. 3F-G). These findings clearly indicate that LssaCA plays an important role in promoting RSV infection of plants after RSV has been inoculated into the plant.

LssaCA interacts with rice thaumatin-like protein to increase its endo-β-1,3-glucanase activity

Because LssaCA played its role in planta, we performed two experiments to investigate the mechanisms by which LssaCA enhances RSV infection. The first was to create LssaCA mutant proteins lacking enzymatic activity to determine if the enhanced infection was related to the LssaCA enzymatic activity. The second was to conduct a yeast two-hybrid screening to identify rice proteins that interact with LssaCA.

LssaCA has 7 residues constituting the enzymatic active center, including H111, N139, H141, H143, E153, H166, and T253. Seven recombinant mutant proteins were constructed and were labeled as LssaCAH111D, LssaCAN139H, LssaCAH141D, LssaCAH143D, LssaCAE153H, LssaCAH166D, LssaCAT253E (Fig. S2A). Enzymatic activity assays indicated that five of the seven mutants had significantly reduced enzymatic activity (Fig. S2B). To measure the influence of these mutants on RSV infection, they were pre-cultured with purified RSV particle solution and were then microinjected into plant leaves. The RSV titers were measured using RT-qPCR at 14 dpi. The results showed that, compared to LssaCA, none of the mutants decreased the protein function in enhancing RSV infection (Fig. S2C). These results suggest that LssaCA enables RSV infection in an enzymatic activity-independent manner.

The yeast two-hybrid screening identified a rice thaumatin-like protein (XP_025879846.1), designated as Oryza sativa thaumatin-like protein (OsTLP) (Fig. 4A-B and S3). The open reading frame of OsTLP encoded for a polypeptide of 327 amino acids and a thaumatin family (THM) domain from residues 27 to 267. A BLAST algorithm-based analysis against sequence databases revealed the sequence identities to TLP 1b. The N-terminal 24 amino acids constituted a secretion signal peptide, suggesting OsTLP as a secreted protein. Pull-down and MST assays were performed to confirm the molecular interaction between OsTLP and LssaCA. The pull-down assays showed that OsTLP and LssaCA had a specific interaction (Fig. 4C). The MST assay revealed a KD of 1.3 ± 2.2 μM for the OsTLP-LssaCA interaction, further confirming the molecular interaction (Fig. 4D).

The interaction between OsTLP and LssaCA plays a role in regulating the enzymatic activity of OsTLP.

A. Schematic diagram of the OsTLP protein sequence. SP, signal peptide; THM, conserved thaumatin family protein sequence. Glyco_hydro_64 domain, glycoside hydrolases of family 64.

B. Yeast two-hybrid assay showing the interaction between OsTLP and LssaCA. SD, synthetically defined medium; Leu, Leucine; Trp, Tryptophan; His, Histidine; 3’AT, 3-amino-1,2,4-triazole; AD, transcription activation domain; BD, DNA-binding domain.

C. Pull-down assays showing the interactions between LssaCA (LssaCA-His) and OsTLP (MBP-OsTLP). MBP was used as a negative control. Both anti-His and anti-MBP antibodies were used to detect corresponding proteins.

D. MST assay showing the interaction between LssaCA (LssaCA-His) and OsTLP (MBP-OsTLP). MBP was used as a negative control. Bars represent SE.

E. The endo-β-1,3-glucanase activity of the purified recombinant OsTLP protein (expressed as MBP-fusion protein). MBP was used to normalize results.

F. The endo-β-1,3-glucanase activity of OsTLP, which was overexpressed in transgenic plants (OsTLP OE) or wild-type plants (WT).

G. Regulation of OsTLP enzymatic activity by LssaCA. ****, p < 0.0001.

Previous studies have reported that TLP orthologs in apples, cherries, tomatoes, barley, etc. have endo-β-1,3-glucanase activity, which works on degrading callose (de Jesus-Pires et al., 2020). To confirm the endo-β-1,3-glucanase activity of OsTLP, an enzyme activity experiment was conducted with the recombinantly expressed OsTLP protein. The results revealed an average endo-β-1,3-glucanase activity of 60 units/mg OsTLP (Fig. 4E). To determine the in-planta activity of OsTLP, OsTLP overexpression transgenic plants were constructed. Enzymatic activity assays confirmed that the transgenic plants had significantly increased β-1,3-glucanase activity (Fig. 4F).

To investigate the influence of the molecular interaction between LssaCA and OsTLP on OsTLP activity, OsTLP was pre-incubated with either LssaCA or a BSA control for 2 h before the enzymatic activity was measured. The results showed that pre-incubation with LssaCA significantly increased OsTLP activity, compared to the BSA control (Fig. 4G). These results imply that LssaCA may facilitate RSV infection by enhancing the activity of OsTLP to degrade callose.

LssaCA inhibits callose deposition induced by RSV infection

Callose deposition on sieve plates is an important mechanism for slowing virus long-distance transmission and reducing the virus infection level in plants (Li et al., 2012; Zavaliev et al., 2011). To investigate whether RSV infection induces callose deposition, rice seedlings were fed by RSV-free or viruliferous L. striatellus for 24 h, and then callose deposition of the plants was measured. The results showed that VIF led to more callose deposition compared with that induced by feeding with RSV-free L. striatellus (Fig. 5A-B). ELISA assays also revealed an increased callose concentration in the plant fed by viruliferous L. striatellus (Fig. 5C). To further confirm the role of RSV in inducing callose deposition, the transcript levels of callose synthase genes were measured using RT-qPCR. We found that RSV infection upregulated gsl3, gsl4, gsl5, and gsl10 (Fig. 5D–G). These results demonstrate that RSV infection induces callose deposition.

LssaCA inhibits RSV-induced callose deposition.

A. Bright blue fluorescence of cross-sections showing callose deposition at feeding sites. Samples were prepared from the leaf phloem of plants fed on by RSV-free or RSV-infected L. striatellus. Thin sections were stained with 0.1% aniline blue at 24 h after L. striatellus feeding and examined under a fluorescence microscope. Xy, xylem; ph, phloem. Scale bars: 20 μm.

B. Average fluorescence intensity of arbitrary area of callose deposition counted using ImageJ. Eight to ten random sites per sample were selected for the evaluation of fluorescence intensity. ****, p < 0.0001.

C. Callose concentration in leaves of rice plants fed on by RSV-free or RSV-infected L. striatellus. ****, p < 0.0001.

D–G. Transcript levels of callose synthase genes as determined by RT-qPCR. Insects were allowed to feed on rice plants for 24 h before total RNAs were extracted. *, p < 0.05; **, p < 0.01.

A. H. Bright blue fluorescence of cross-sections showing callose deposition at feeding sites. Samples were prepared from leaf phloem of plants fed on by dsGFP- or dsLssaCA-treated RSV-infected L. striatellus. Thin sections were stained with 0.1% aniline blue at 24 h after L. striatellus feeding and examined under a fluorescence microscope. Xy, xylem; ph, phloem. Scale bars: 20 μm.

I. Average fluorescence intensity of arbitrary area of callose deposition counted using ImageJ under a fluorescence microscope. Eight to ten random sites per sample were selected for the evaluation of fluorescence intensity. **, p < 0.01.

B. J. Callose concentration in leaves of rice plants fed on by dsGFP- or dsLssaCA-treated L. striatellus. **, p < 0.01.

To determine if LssaCA influences callose deposition induced by RSV infection, we downregulated LssaCA in viruliferous L. striatellus by dsRNA microinjection. On 2 dpi when LssaCA expression was significantly reduced, the insects were allowed to feed on healthy plants for 24 h, and then the callose deposition was measured. The results showed that the plants fed by LssaCA-deficient insects deposited more callose compared to the plants fed by the control-group insects (dsGFP treatment) (Fig. 5H–J), indicating that LssaCA can impair the callose caused by RSV infection.

OsTLP degrades callose to facilitate RSV infection

To confirm if OsTLP degrades the RSV-induced callose in planta, viruliferous L. striatellus were allowed to feed on either OsTLP overexpression plants or control plants, and then callose deposition was measured. The results showed that transgenic plants exhibited significantly lower callose deposition compared to wild-type ones (Fig. 6A-B), indicating that OsTLP can degrade the deposit callose induced by RSV infection. Moreover, we measured the RSV infection levels that were influenced by OsTLP overexpression. Viruliferous L. striatellus fed on either OsTLP overexpression plants or control plants for 2 days, and then the RSV titers were measured on 14 dpf. The RT-qPCR analyses indicated that, compared with wild-type plants, plants overexpressing OsTLP had significantly higher RSV titers (Fig. 6C), indicating that OsTLP facilitates RSV infection in planta. It is hypothesized that the overexpressed OsTLP degrades the deposited callose that is caused by RSV infection, therefore facilitating RSV infection.

OsTLP inhibits callose deposition to facilitate RSV infection.

A. Bright blue fluorescence of cross-sections showing callose deposition at feeding sites. Samples were prepared from the leaf phloem of plants fed on by RSV-infected L. striatellus. Scale bars: 20 μm. OsTLP OE, transgenic plants over-expressing OsTLP.

B. Average fluorescence intensity of arbitrary area of callose deposition counted using ImageJ under a fluorescence microscope. *, p < 0.05.

C. Promotion of RSV infection by OsTLP overexpression, as determined by RT-qPCR. Titers of RSV (copy number of NP) were measured at 14 days post-feeding. *, p < 0.05.

RSV NP interacts with LssaCA to further enhance OsTLP enzymatic activity

We have confirmed that LssaCA binds to OsTLP to promote callose degradation and facilitates RSV infection in planta, we then investigated whether there is a direct interaction between LssaCA and RSV by which RSV directly participates in the function played by LssaCA. By using RSV and LssaCA-specific antibodies to conduct immunofluorescence assays (IFA), we found that LssaCA and RSV co-localized in salivary glands (Fig. 7A), suggesting a potential molecular interaction between them. A pull-down assay then revealed a specific interaction between LssaCA and RSV NP (Fig. 7B). An MST assay determined that the KD of the interaction between LssaCA and NP was 2.7 ± 2.2 μM, further confirming the molecular interaction (Fig. 7C).

LssaCA-RSV NP interaction enhances the OsTLP enzymatic activity.

A. Immunofluorescence assay showing co-localization of RSV (shown in red) and LssaCA (shown in green) in L. striatellus salivary glands. Images are representative of three independent experiments with a total of 30 SBPHs analyzed. The scale bar represents 20 μm. psg, principal salivary gland; asg, accessory salivary gland.

B. GST pull-down assays show the interaction between LssaCA (GST-LssaCA) and RSV NP (NP-His). GST was used as a negative control. Anti-GST and anti-His antibodies were used to detect the corresponding proteins.

C. MST assay showing the interaction between LssaCA (LssaCA-His) and RSV NP (GST-NP). GST was used as a negative control. Bars represent SE.

D. Pull-down assays showing the interaction between OsTLP (MBP-OsTLP) and LssaCA-NP complex. LssaCA was expressed with a His tag and RSV NP was expressed as a GST-fusion protein. LssaCA and NP were preincubated before co-incubation with OsTLP. Anti-His and anti-GST antibodies were used to detect the corresponding proteins.

A. E. Regulation of OsTLP enzymatic activity by LssaCA and LssaCA-NP. (LssaCA + NP) indicates that the two proteins were pre-incubated before incubation with OsTLP. ****, p < 0.0001.

As LssaCA interacts with both RSV NP and OsTLP, we measured the tripartite interaction. Through pull-down assays, we discovered that when LssaCA was pre-incubated with RSV to form an NP-LssaCA complex, the complex still had the ability to bind to OsTLP (Fig 7D). This result demonstrates a tripartite interaction among these three proteins.

We then explored how the tripartite interaction affects OsTLP enzymatic activity. OsTLP was pre-incubated with LssaCA-NP complex or LssaCA control. After incubation for 2 h, the enzyme activity was measured. The results showed that pre-incubation with LssaCA-NP complex significantly increased OsTLP activity compared with LssaCA control (Fig. 7E), indicating that stronger activation of β-1,3-glucanase activity was achieved through the NP-LssaCA-OsTLP tripartite interaction.

Taken together, these results suggested that when RSV is delivered into plant phloem, it can induce callose deposition. This process is overturned through a tripartite interaction involving the LssaCA protein which binds to RSV in insect salivary glands and then the LssaCA-RSV complex binds to OsTLP in plants. Once this tripartite interaction occurs, OsTLP activity of β-1,3-glucanase is significantly increased which degrades callose thus allowing for efficient RSV infection.

Discussion

Insects are known to be vectors of many plant viruses, and some of these plant viruses rely entirely on insect vectors for their plant-to-plant transmission. Previous studies have suggested that insect salivary proteins may facilitate arbovirus infection of plants through various mechanisms, such as breaking down plant cell walls, inhibiting plant immune defenses and callose deposition, directly binding to arbovirus to prevent their detection by plant immune systems, and more. Further research is needed to fully understand the role that insect salivary proteins play in arboviruses infection of plants. In this study, we have found that RSV infection of rice plants triggers the deposition of callose. However, this process is reversed by a tripartite interaction involving the central component of an insect salivary protein LssaCA, which subsequently binds to RSV in insect salivary glands and then OsTLP in the plants. The activity of OsTLP’s β-1,3-glucanase is significantly increased when this tripartite interaction occurs; leading to the degradation of callose and thus allowing for efficient RSV infection (Fig. 8).

Proposed model of how RSV NP-LssaCA-OsTLP tripartite interaction leads to callose degradation to promote RSV infection of plants.

A. L. striatellus feeding on rice plants inoculates both RSV and LssaCA into the phloem sieve elements. Within the plants, RSV induces callose synthesis, causing callose deposition to inhibit RSV infection. However, LssaCA reverses this inhibition by binding to OsTLP to enhance its β-1,3-glucanase activity, which degrades the deposited callose. The interaction between RSV and LssaCA further enhances the OsTLP enzymatic activity and facilitates RSV infection.

Callose is a complex polysaccharide, and its deposition in plant phloem plays an important role in defending against insect attacks and viral spread (Zavaliev et al., 2011). When attacked by piercing insects or viruses, plants activate callose deposition on the sieve plates to occlude the flow of phloem sap, thus discouraging feeding and reducing viral long-distance spread. This accumulation of callose in plant phloem can be switched on or off through the action of callose synthases and hydrolases (Li et al., 2012; Shangguan et al., 2018). Studies have shown that expression levels of these glucanase enzymes differ between resistant and susceptible plant species. This suggests that some plants may be more successful at resisting infection due to high levels of glucanase enzymes in their tissues, which allow them to more quickly and effectively form callose barriers against insects and viruses (Hao et al., 2008). Some insects have evolved β-1,3-glucanases enzymes that can degrade the deposited callose, thus enabling them to bypass the plant’s defense mechanisms and continue feeding (Bucher et al., 2001; Hao et al., 2008). In this study, we demonstrate a novel mechanism by which an insect saliva protein binds to a plant β-1,3-glucanases protein, thus activating the enzymatic activity and promoting callose degradation.

In this study, we found that plants fed on by viruliferous insects showed higher callose deposition than did plants fed on by uninfected insects (Fig. 5A–C), suggesting that arbovirus infection via insect saliva may represent a double stress to the plant, inducing stronger callose deposition that must be overcome for successful insect feeding and viral spread. To analyze how this happens, we looked at the tripartite interaction between LssaCA-NP-OsTLP. We found that this interaction induced significantly higher β-1,3-glucanase activity than the LssaCA-OsTLP interaction alone, suggesting that SBPH and RSV have evolved mechanisms to increase β-1,3-glucanase activity in order to bypass the plant’s defense of callose deposition and facilitate successful viral spread.

LssaCA is carbonic anhydrase, but its role in aiding RSV infection does not rely on its enzymatic activity (Fig. S2). Our results found that even though LssaCA mutants had significantly reduced enzymatic activity, their ability to facilitate RSV infection was not significantly decreased. Because LssaCA interacts with RSV NP, we measured the interactions between LssaCA mutants and NP. Pull-down assays indicated that all the mutants still interacted with RSV NP (Fig. S4), while the mutants with a stronger interaction (LssaCAH111D and LssaCAE153H) showed better facilitation for RSV infection levels (Fig. S2). These results further supported the importance of LssaCA-NP interaction in mediating RSV infection, while the exact role of LssaCA enzymatic activities in aiding RSV infection is still unclear.

In conclusion, this study elucidated a tripartite interaction among a plant virus, an insect saliva protein, and a host phloem β-1,3-glucanase. This interaction inhibited the plant defense of callose deposition and enabled arboviral phloem transmission. This study provides new insights into the tripartite virus-insect vector-plant interaction, which is relevant to many agriculturally important plant arboviruses whose transmission is facilitated via insect saliva proteins.

Figure S1. Nucleotide sequence and deduced amino acid sequence of LssaCA.

Underlining indicates the predicted signal peptide sequence. Seven predicted catalytically active residues are shown in red. Bold font indicates eukaryotic-type carbonic anhydrase sequences.

Figure S2. LssaCA enzymatic activity does not affect RSV infection.

A. Summary of the predicted enzymatic active sites and the muteins.

B. Relative esterase activities of recombinant LssaCA protein and its muteins. *, p < 0.05; **, p < 0.01; ****, p < 0.0001.

C. Schematic diagram to show the process of microinjecting RSV into a plant. The RSV was pre-incubated with recombinantly expressed LssaCA, the LssaCA muteins, or a BSA control, before being microinjected into the rice phloem.

D. The effects of LssaCA, its muteins, or the BSA control, on the levels of RSV infection as determined by RT-PCR. The titers of RSV (copy number of NP) were measured at 14 dpi.

Figure S3. Nucleotide sequence and deduced amino acid sequence of OsTLP.

Underlining indicates the predicted signal peptide sequence. Bold font indicates the conserved thaumatin family protein sequence.

Figure S4. The molecular interaction between RSV and LssaCA muteins is important for RSV to infect rice plants.

A. Pull-down assays showing the interactions between the LssaCA muteins (expressed with a His-tag) and RSV NP (GST-NP). GST was used as a negative control. Anti-His and Anti-GST represent antibodies against 6*His and GST, respectively. Both LssaCAH111D and LssaCAE153H (marked in red) have a stronger interaction with NP compared to the wild-type LssaCA protein.

B. Preincubation of RSV with either LssaCAH111D or LssaCAE153H facilitates RSV infection of plants after being delivered via microinjection into the phloem. The titers of RSV (copy number of NP) were measured at 14 dpi. BSA was used as a negative control. *, p < 0.05; **, p < 0.01.

Table S1. Primers used in this study.

Materials and methods

Viruses, insects, host plants, and antibodies

The RSV-free and RSV-infected L. striatellus individuals used in this study were originally captured in Jiangsu Province, China, and were maintained in our laboratory. All plants used to rear L. striatellus were grown in a growth incubator (14.5 cm height, 3.2 cm radius) at 25℃ under a 16-h light/8-h dark photoperiod. Insects were transferred to fresh seedlings every 15 days to ensure adequate nutrition and insect activity.

The RSV-specific antibody was produced using RSV ribonucleoproteins (RNPs) as antigens. The LssaCA polyclonal antibody was produced using recombinantly expressed LssaCA protein (with a His-tag at the C-terminus) as the antigen.

Tissue collection

Insect dissection and tissue collection were performed using previously described procedures (Huo et al., 2018). In brief, insects were anesthetized at 4°C for 10 min, and the forelegs were severed at the coxa-trochanter joint using forceps. The hemolymph was expelled and drawn to the tip of clean forceps. The insects were then dissected from the abdomen in pre-chilled PBS buffer. The fat body was collected with forceps, aided by the tension of the liquid. Tissues, including the guts, salivary glands, and ovaries, were washed three times in PBS to remove contaminants from the hemolymph.

Bioinformatic analysis

The signal peptide and transmembrane helix were predicted using tools at the SignalP 5.0 server (https://services.healthtech.dtu.dk/service.php?SignalP-5.0) and the TMHMM-2.0 server (https://services.healthtech.dtu.dk/service.php?TMHMM-2.0). Homology searches of the protein sequence of LssaCA and OsTLP were conducted using the BLAST algorithm at the National Center for Biotechnology Information (http://blast.ncbi.nlm.nih.gov/Blast.cgi). Conserved domains were predicted with SMART (http://smart.embl-heidelberg.de/).

Collection of watery and gel saliva

Saliva was collected as described previously (Huo et al., 2022). Watery saliva was obtained by centrifuging the collected artificial diet at 10,000 ×g for 10 min to remove the solid gel. To collect gel saliva proteins, the membrane was washed twice with PBS. Sheathes adhering to the parafilm were denatured and solubilized in a solution consisting of 8 M urea, 4% v/v CHAPS, 0.1% w/v sodium dodecyl sulfate (SDS), and 2% v/v DTT. The parafilm was incubated on an orbital shaker at room temperature for 45 min. Then, the denatured gel saliva was concentrated to a final volume of 100 µL. Saliva samples were collected from 2,000 insects for silver staining and western blotting analyses.

Measurement of carbonic anhydrase activity

Carbonic anhydrase activity was characterized by measuring its esterase activity. P-nitroethyl acetate (p-NPA) was used as the substrate. The LssaCA enzyme solution was prepared as follows: LssaCA was expressed in E. coli with a His-tag at the C-terminus and the purified LssaCA protein was diluted in PBS buffer (pH 7.0) to a final concentration of 1 mg/mL. Details of the reaction conditions have been described elsewhere (Verpoorte et al., 1967). Briefly, 1 mL of freshly prepared 3 mM p-NPA dissolved in acetone was added to 1.9 mL 100 mM PBS (pH 7.0), followed by the addition of 0.1 ml LssaCA enzyme solution (0.5 mg/mL in 100 mM PBS). As the blank control, 2 mL 100 mM PBS (pH 7.0) was added to 1 mL 3 mM p-NPA. The absorbance at 384 nm was measured at the start of the reaction and 3 min later. One unit (U) of enzyme activity is defined as the amount of enzyme that hydrolyzes 1 μmol p-nitroethyl acetate per minute at pH 7.0 at 25 °C.

Immunofluorescence assays

Immunofluorescence assays (IFA) were performed to determine the localization of RSV and LssaCA in L. striatellus salivary glands. The dissected salivary gland samples were fixed in 4% v/v paraformaldehyde at 4°C overnight and then permeabilized in 2% v/v Triton X-100 for 2 h. The samples were incubated with rabbit anti-RSV serum and mouse anti-LssaCA serum (1:250 dilution in PBST/FBS: PBS containing 2% v/v Triton X-100 and 2% fetal bovine serum) for 1 h at room temperature. After washing twice with PBST, the samples were incubated with 1:200 diluted Alexa Fluor 488-conjugated goat anti-mouse antibody and Alexa Fluor 594-conjugated goat anti-rabbit antibody (Invitrogen) for 1 h. After washing three times with PBST, the samples were examined under a Leica TCS SP8 microscope (Leica, Wetzlar, Germany).

Evaluation of callose deposition in rice plants

Five insects were allowed to feed on a 10 d-old rice plant in a leaf area of 2 × 2 cm for 24 h. The leaf samples were immersed in 5 mL alcoholic lactophenol (1 volume phenol: glycerol: lactic acid: water (1:1:1:1) and 2 volumes ethanol), then evacuated with a syringe for 15 min, and then incubated at 65℃ for 30 min until the chlorophyll was completely cleared. Cleared leaf samples were rinsed with 50% v/v ethanol, then with water, and then stained in 150 mM K2HPO4 (pH 9.5) containing 0.01% w/v aniline blue for 30 min in the dark. Stained leaf samples were sectioned and mounted in 50% v/v glycerol on slides. The slides were examined under a Zeiss observer Z1 immunofluorescence microscope (Carl Zeiss MicroImaging GmbH, Gottingen, Germany), and fluorescence intensity was determined using ImageJ software.

Callose concentrations were measured using a plant callose ELISA kit (Catalog number: JL46407; Jonln, Shanghai, China), according to the manufacturer’s protocol. The callose solution was prepared as follows: leaf material from the feeding site was homogenized with PBS buffer (pH 7.4) and the mixture was centrifuged at 500–1000 g for 20 min. The supernatant was collected for the detection of callose concentration. The absorbance at 450 nm was measured and callose concentrations were calculated based on the standard curve.

RNA interference

We conducted RNAi analyses to determine the function of LssaCA in mediating RSV infection and regulating callose deposition. The LssaCA-specific gene fragment was PCR-amplified with the primer pair LssaCA-T7F/LssaCA-T7R (Table S1). Then, dsRNA was synthesized using a commercial kit (T7 RiboMAX Express RNAi System, Promega, Madison, WI, USA) and purified by phenol: chloroform extraction and isopropanol precipitation. Finally, 36.8 nL dsRNA at 1 ng/nL was delivered into the insect hemocoel for gene silencing. Microinjection was performed using a glass needle through a Nanoliter 2000 (World Precision Instruments, Sarasota, FL, USA). The negative control, GFP dsRNA, was synthesized and microinjected following the same protocol.

RT-qPCR

Gene transcript levels were determined by RT-qPCR. The RNA was extracted from L. striatellus tissues using an RNeasy Mini Kit (Qiagen, Hilden, Germany), and from plant leaves using TRIzol. Reverse transcriptional PCR (iScript cDNA Synthesis Kit, Bio-Rad, Hercules, CA, USA) and SYBR-Green-based qPCR (SYBR Green Supermix, Bio-Rad) was performed according to the manufacturer’s protocols. The primer pair used to amplify LssaCA was LssaCA-q-F/LssaCA-q-R (Table S1). Viral RNA copies were measured using the primers NP-q-F/NP-q-R (Table S1). L. striatellus translation elongation factor 2 (EF2) and rice actin were amplified as internal controls to ensure equal loading of cDNA isolated from different samples. The primer pairs EF2-q-F/EF2-q-R and actin-q-F/actin-q-R (Table S1) were used to amplify EF2 and actin, respectively. Water was used as a negative control.

Yeast two-hybrid assays

Yeast two-hybrid screening was performed to identify rice proteins interacting with LssaCA. A high-complexity rice cDNA library was fused to Gal4 AD and transformed into the yeast strain Y187. The LssaCA gene lacking the signal-peptide sequence (LssaCAsp-) was fused to Gal4 DNA-BD and transformed into the yeast strain Y2HGold. The two yeast strains were co-cultured overnight to allow mating to create diploids. After library screening, positive clones were selected on quadruple dropout medium: SD/–Ade/–His/–Leu/–Trp supplemented with X-a-Gal and aureobasidin A (QDO/X/A), and prey plasmids were isolated for sequencing. To further confirm the interaction between LssaCA and OsTLP, LssaCAsp- was cloned into the prey vector pDB-Leu, and the OsTLP gene lacking the signal-peptide sequence (OsTLPsp-) was cloned into the bait vector pPC86. The two plasmids were co-transformed into the yeast stain Mav203 and the transformants were selected on SD/–His–Leu–Trp + 20 mM 3-AT medium.

Pull-down assays

Pull-down assays were performed to determine the molecular interactions between LssaCA and RSV NP, LssaCA muteins and RSV NP, LssaCA and OsTLP, and OsTLP and the LssaCA-NP complex.

The interaction between LssaCA and RSV NP was determined using GST pull-down analyses as described previously (Huo et al., 2014). Briefly, LssaCAsp- was cloned into the pGEX-3x vector and LssaCA was expressed as a GST-fusion protein (GST-LssaCA). The NP gene was cloned into the pET-30a (+) expression vector and the protein was expressed with an N-terminal His-tag (NP-His). The GST-LssaCA construct was expressed in, and purified from, E. coli stain BL21. The GST-LssaCA protein or the GST control protein was bound to Glutathione Sepharose beads (Cytiva, Marlborough, MA, USA), then NP-His protein was added to the beads. The bead-bound proteins were separated by SDS-PAGE and detected by western blotting with antibodies against GST or His-tag.

We conducted His pull-down assays to determine the interactions between the LssaCA muteins and RSV NP. The gene sequences encoding LssaCA muteins (without the signal-peptide sequence) were each cloned into the pET-30a (+) expression vector, and each mutein was expressed with an N-terminal His-tag. The NP gene was cloned into the pGEX-3x vector and NP was expressed as a GST-fusion protein (GST-NP). Each of the His-tag-fused LssaCA muteins was bound to Ni Sepharose (GE Healthcare, Uppsala, Sweden), then the GST-NP protein or the GST control protein was added to the beads. The bead-bound proteins were separated by SDS-PAGE and detected by western blotting with antibodies against GST or His-tag.

Next, MBP pull-down analyses were performed to determine the interaction between OsTLP and LssaCA. The OsTLPsp- sequence was cloned into pMAL-p2X to produce the MBP-fusion protein (MBP-OsTLP). The MBP-OsTLP protein or the MBP control protein was bound to amylose resin (New England Biolabs, Beverley, MA, USA), and then the LssaCA-His protein was added to the beads. To determine the interaction between OsTLP and the LssaCA-NP complex, the LssaCA-His and GST-NP proteins were preincubated before being added to the beads. The bead-bound proteins were separated by SDS-PAGE and detected by western blotting with the MBP antibody (Cell Signaling Technology, Danvers, MA, USA), His-tag antibody (Tiangen, Beijing, China) and GST antibody (Tiangen).

Microscale thermophoresis (MST) assays

We conducted MST assays to detect the interactions between LssaCA and NP, and between LssaCA and OsTLP. First, 10 μM purified LssaCAsp- protein (His-fusion protein) was labelled with a Monolith NT Protein Labeling Kit RED-NHS (Nano Temper Technologies GMBH, München, Germany) using red fluorescent dye NT-647 N-hydroxysuccinimide (amine-reactive), according to the manufacturer’s instructions. The binding assays were performed on a Monolith NT.115 Microscale Thermophoresis instrument (Nano Temper Technologies GMBH) using capillaries treated using the standard method. The labeled protein LssaCAsp- was added to serially diluted NP (GST-NP, GST was used as a negative control) or OsTLPsp- (MBP-fusion protein, MBP was used as a negative control). The initial concentration of each protein was 40 µM with 0.1% (v/v) Tween 20. The KD Fit function of NanoTemper Analysis software version 1.2.214.1 was used for curve fitting and to calculate the dissociation constant (Kd).

RSV inoculation through midrib microinjection

Midrib microinjection was performed as described previously (Zhao et al., 2017). Briefly, 200 RSV-infected L. striatellus were homogenized with 200 μL PBS (pH 7.0), and the homogenate was centrifuged twice at 8000 g for 10 min at 4℃. The crude virus extract was transferred to a new centrifuge tube. The concentration of RSV in the extract was measured using a Nanodrop One instrument. The RSV extract was adjusted to 1 mg/mL for microinjection. Rice seedlings were grown in an incubator for 2 weeks until the leaf veins became clearly visible. Then, 23 nL RSV extract was microinjected into the midrib. The RSV titers in plants were detected at 7–14 days post microinjection.

β-1,3-glucanase activity

β-1,3-glucanase hydrolyzes laminarin and cuts the β-1,3-glucoside bond to produce a reducing terminus. Therefore, the rate of reducing sugars production was used to calculate enzyme activity. One unit of β-1,3-glucanase activity was defined as the amount of enzyme that hydrolyses laminarin to generate 1 μg reducing sugars per minute.

The β-1,3-glucanase activity of OsTLP was measured using a beta-1,3-glucanase microplate assay kit (Cat. ASK1028, Bioworld, Bloomington, MN, USA) according to the manufacturer’s instructions.

To determine the enzymatic activity of the recombinant OsTLP proteins, the MBP-fusion OsTLPsp- protein or the MBP control protein was adjusted to 0.5 mg/mL. To determine the enzymatic activity of the in vivo-expressed OsTLPsp-, 0.1 g rice leaf material was homogenized with 1 mL assay buffer on ice, and then the mixture was centrifuged at 12,000g at 4°C for 10 min. The supernatant was transferred into a new centrifuge tube and kept on ice until detection of enzymatic activity.

For the assay, a 50-μL aliquot of the protein solution was mixed with laminarin substrate and incubated at 37°C for 30 min. The reaction was terminated by incubating the reaction tube in a boiling water bath for 10 min. The reaction supernatant was transferred into a microplate and incubated at 90°C for 10 min before measuring the absorbance at 540 nm.

Generation of transgenic plants

The full-length OsTLP gene was PCR-amplified with the primer pair TLP-F/TLP-R (Table S1), and cloned into the plant expression vector pCAMBIA-1300-eGFP to produce the plasmid pCAMBIA-1300-OsTLP. The plasmid pCAMBIA-1300-OsTLP was transformed into seedlings of the Japonica rice cultivar Nipponbare through Agrobacterium-mediated transformation as described previously (Hiei et al., 1994). Thirty independent T0 transformants were obtained and were screened by germinating seeds from T2 lines on medium containing 50 μg/ml hygromycin. The hygromycin-resistant homozygous seeds were used for further experiments. Seeds from homozygous lines transformed with the empty vector pCAMBIA-1300-eGFP were used as the negative control.

Statistical analyses

Graphs were constructed and statistical analyses were performed using Prism 9.0 software (GraphPad Software, San Diego, CA, USA). Data are expressed as the mean ± standard deviation (SD) calculated from three independent experiments. The significance of differences between two groups was determined by unpaired Student’s t-test. Differences among three or more groups were compared using one-way analysis of variance (ANOVA).

eLife Assessment

This study presents a useful set of experiments to test how a salivary protein might facilitate planthopper-transmitted rice stripe virus infection by interfering with callose deposition. If the conclusions can be confirmed, the study will enhance our understanding of the mechanisms mediating tripartite virus-insect vector-plant interactions and would be of general interest in plant science research. Whereas most experimental data is compelling, the conclusions rely on inadequate evidence for the salivary protein carbonic anhydrase being present two weeks post injection in the plant, which seems unlikely.

Reviewer #1 (Public Review):

In this study, the authors identify an insect salivary protein participating viral initiate infection in plant host. They found a salivary LssaCA promoting RSV infection by interacting with OsTLP that could degrade callose in plants. Furthermore, RSV NP bond to LssaCA in salivary glands to form a complex, which then bond to OsTLP to promote degradation of callose.

The story focus on tripartite virus-insect vector-plant interaction, and is interesting. However, the study is too simple and poor-conducted. The conclusion is also overstated due to unsolid findings.

Major comments:

1. The key problem is that how long the LssCA functioned for in rice plant. Author declared that LssCA had no effect on viral initial infection, but on infection after viral inoculation. It is unreasonable to conclude that LssCA promoted viral infection based on the data that insect inoculated plant just for 2 days, but viral titer could be increased at 14 day post-feeding. How could saliva proteins, which reached phloem 12-14 days before, induce enough TLP to degrade callose to promote virus infection? It was unbelievable.

2. Lines 110-116 and Fig. 1, the results of viruliferous insect feeding and microinjection with purified virus could not conclude the saliva factor necessary of RSV infection, because these two tests are not in parallel and comparable. Microinjection with salivary proteins combined with purified virus is comparable with microinjection with purified virus.

The second problem is how many days post viruliferous insect feeding and microinjection with purified virus did author detect viral titers? in Method section, authors declared that viral titers was detected at 7-14 days post microinjection. Please demonstrate the days exactly.

The last problem is that how author made sure that the viral titers in salivary glands of insects between two experiments was equal, causing different phenotype of rice plant. If not, different viral titers in salivary glands of insects between two experiments of course caused different phenotype of rice plant.

3. The callose deposition in phloem can be induced by insect feeding. In Fig. 5H, why was the callose deposition increased in the whole vascular bundle, but not phloem? Could the transgenic rice plant directional express protein in the phloem? In Fig.5, why was callose deposition detected at 24 h after insect feeding? In Fig. 6A, why was callose deposition decreased in the phloem, but not all the cells of the of TLP OE plant? Also in Fig.6A and B, expression of callose synthase genes was required.

Reviewer #2 (Public Review):

There is increasing evidence that viruses manipulate vectors and hosts to facilitate transmission. For arthropods, saliva plays an essential role for successful feeding on a host and consequently for arthropod-borne viruses that are transmitted during arthropod feeding on new hosts. This is so because saliva constitutes the interaction interface between arthropod and host and contains many enzymes and effectors that allow feeding on a compatible host by neutralizing host defenses. Therefore, it is not surprising that viruses change saliva composition or use saliva proteins to provoke altered vector-host interactions that are favorable for virus transmission. However, detailed mechanistic analyses are scarce. Here, Zhao and coworkers study transmission of rice stripe virus (RSV) by the planthopper Laodelphax striatellus. RSV infects plants as well as the vector, accumulates in salivary glands and is injected together with saliva into a new host during vector feeding.

The authors present evidence that a saliva-contained enzyme - carbonic anhydrase (CA) - might facilitate virus infection of rice by interfering with callose deposition, a plant defense response. In vitro pull-down experiments, yeast two hybrid assay and binding affinity assays show convincingly interaction between CA and a plant thaumatin-like protein (TLP) that degrades callose. Similar experiments show that CA and TLP interact with the RSV nuclear capsid protein NT to form a complex. Formation of the CA-TLP complex increases TLP activity by roughly 30% and integration of NT increases TLP activity further. This correlates with lower callose content in RSV-infected plants and higher virus titer. Further, silencing CA in vectors decreases virus titers in infected plants. Interestingly, aphid CA was found to play a role in plant infection with two non-persistent non-circulative viruses, turnip mosaic virus and cucumber mosaic virus (Guo et al. 2023 doi.org/10.1073/pnas.2222040120), but the proposed mode of action is entirely different.

While this is an interesting work, there are, in my opinion, some weak points. The microinjection experiments result in much lower virus accumulation in rice than infection by vector inoculation, so their interpretation is difficult. Also, the effect of injected recombinant CA protein might fade over time because of degradation or dilution. The authors claim that enzymatic activity of CA is not required for its proviral activity. However, this is difficult to assess because all CA mutants used for the corresponding experiments possess residual activity. It remains also unclear whether viral infection deregulates CA expression in planthoppers and TLP expression in plants. However, increased CA and TLP levels could alone contribute to reduced callose deposition.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

In this study, the authors identify an insect salivary protein participating viral initiate infection in plant host. They found a salivary LssaCA promoting RSV infection by interacting with OsTLP that could degrade callose in plants. Furthermore, RSV NP bond to LssaCA in salivary glands to form a complex, which then bond to OsTLP to promote degradation of callose.

The story focus on tripartite virus-insect vector-plant interaction and is interesting. However, the study is too simple and poor-conducted. The conclusion is also overstated due to unsolid findings.

We thank the reviewer for their constructive feedback. We have conducted additional experiments to strengthen our results and conclusions as detailed below:

(1) The comparison between vector inoculation and microinjection involves multiple confounding factors that could affect the experimental results, including salivary components, RSV inoculation titers, and the precision of viral deposition. The differential outcomes could be attributed to these various factors rather than definitively demonstrating the necessity of salivary factors. Therefore, we have removed this comparison from the revised manuscript and instead focused on elucidating the specific mechanisms by which LssaCA facilitates viral infection.

(2) We conducted new experiments to assess the function of LssaCA enzymatic activity in mediating RSV infection. Additional experiments revealed that OsTLP enzymatic activity is highly pH-dependent, with increased activity as pH decreases from 7.5 to 5.0 (Fig. 3H). However, the LssaCA-OsTLP interaction at pH 7.4 significantly enhanced OsTLP enzymatic activity without requiring pH changes. These results demonstrate that LssaCA-OsTLP protein interactions are crucial for mediating RSV infection. In contrast to pH-dependent mechanisms, our study demonstrated that LssaCA's biological function in mediating RSV infection is at least partially, if not completely, independent of its enzymatic activity. We have added these new resulted into the revised manuscript (Lines 220-227). We have also added a comprehensive discussion comparing the aphid CA mechanism described by Guo et al. (2023 doi.org/10.1073/pnas.2222040120) with our findings in the revised manuscript (Lines 350-371).

(3) We have repeated majority of callose deposition experiments, providing clearer images (Figures 5-6). In addition to aniline blue staining, we quantified callose concentrations using a plant callose ELISA kit to provide more precise measurements (Figure 5A, I, 6A, C and S8A). We utilized RT-qPCR to measure callose synthase expression in both feeding and non-feeding areas, confirming that callose synthesis was induced specifically in feeding regions, leading to localized callose deposition (Figures 5D-G and S8B-E). For sieve plate visualization, we examined longitudinal sections, which revealed callose deposition in sieve plates during SBPH feeding and RSV infection (Figure S7).

(4) We generated OsTLP mutant rice seedlings (ostlp) and use this mutant to directly demonstrate that LssaCA mediates callose degradation in planta through enhancement of OsTLP enzymatic activity (Lines 288-302 and Figure 6).

(5) We produced LssaCA recombinant proteins in sf9 cells to ensure full enzymatic activity and constructed a comprehensive CA mutant protein, in which all seven residues constituting the enzymatic active center mutated (LssaCAH111D,LssaCAN139H,LssaCAH141D, LssaCAH143D, LssaCAE153H, LssaCAH166D, LssaCAT253E) (Fig. S1B). This LssaCA mutant protein demonstrated complete loss of enzymatic activity (Fig. 1C).

Major comments:

(1) The key problem is that how long the LssCA functioned for in rice plant. Author declared that LssCA had no effect on viral initial infection, but on infection after viral inoculation. It is unreasonable to conclude that LssCA promoted viral infection based on the data that insect inoculated plant just for 2 days, but viral titer could be increased at 14 days post-feeding. How could saliva proteins, which reached phloem 12-14 days before, induce enough TLP to degrade callose to promote virus infection? It was unbelievable.

We appreciate your insightful comment and acknowledge that our initial description may have been unclear. We agree that salivary proteins would not present in plant tissues for two weeks post-feeding or post-injection. Our intention was to clarify that when salivary proteins enhance RSV infection, this initial enhancement leads to sustained high viral loads. We measured viral burden at 14 days post-feeding or post-injection because this is the common measurement time point when viral titers are sufficiently high for reliable detection by qRT-PCR or western blotting. We have clarified this rationale in the revised manuscript (Lines 155-157).

To determine the actual persistence of LssaCA in plant tissues, we conducted additional experiments where insects were allowed to feed on a defined aera of rice seedlings for two days. We then monitored LssaCA protein levels at 1 and 3 days after removing the insects. Western blotting analysis revealed that LssaCA protein levels decreased post-feeding and remained detectable at 3 days post-feeding. These results are presented in Figure 2H and described in detail in Lines 184-193.

(2) Lines 110-116 and Fig. 1, the results of viruliferous insect feeding and microinjection with purified virus could not conclude the saliva factor necessary of RSV infection, because these two tests are not in parallel and comparable. Microinjection with salivary proteins combined with purified virus is comparable with microinjection with purified virus.

We thank the reviewer’s insightful comment. We agree that “the results of viruliferous insect feeding and microinjection with the purified virus could not conclude the saliva factor necessary of RSV infection”. However, due to the technical difficulty in collecting sufficient quantities of salivary proteins to conduct the microinjection experiment, we have removed these results from the revised manuscript.

(3) The second problem is how many days post viruliferous insect feeding and microinjection with purified virus did author detect viral titers? in Method section, authors declared that viral titers was detected at 7-14 days post microinjection. Please demonstrate the days exactly.

We thank the reviewer’s insightful comment. We typically measured RSV infection levels at both 7- and 14-days post-microinjection. However, since the midrib microinjection experiments have been removed from the revised manuscript, this methodology has also been removed accordingly.

(4) The last problem is that how author made sure that the viral titers in salivary glands of insects between two experiments was equal, causing different phenotype of rice plant. If not, different viral titers in salivary glands of insects between two experiments of course caused different phenotype of rice plant.

We thank the reviewer’s comment. When we compared the effects of LssaCA deficiency on RSV infection of rice plants, we have compared the viral titers in the insect saliva and salivary glands. The results indicated that the virus titers in both tissues have not changed by LssaCA deficiency, suggesting that the viruses inoculated into rice phloem by insects of different treatments were comparable. Please refer to the revised manuscript Figures 2D-G and Lines 161-173.

(5) The callose deposition in phloem can be induced by insect feeding. In Fig. 5H, why was the callose deposition increased in the whole vascular bundle, but not phloem? Could the transgenic rice plant directional express protein in the phloem? In Fig. 5, why was callose deposition detected at 24 h after insect feeding? In Fig. 6A, why was callose deposition decreased in the phloem, but not all the cells of the of TLP OE plant? Also in Fig.6A and B, expression of callose synthase genes was required.

We thank the reviewer for these insightful comments.

(1) Figure 5. The callose deposition increased in multiple cells within the vascular bundle, including sieve tubes, parenchymatic cells, and companion cells. While callose deposition was detected in other parts of the vascular bundle, no significant differences were observed between treatments in these regions, indicating that in response to RSV infection and other treatments, altered callose deposition mainly occurred in phloem cells. Please refer to the revised 5B, 5J, 6B, and 6D.

(2) Transgenic plant expression. The OsTLP-overexpressing transgenic rice plants express TLP proteins in various cells under the control of CaMV 35S promoter, rather than being directionally expressed in the phloem. However, since TLP proteins are secreted, they are potentially transported and concentrated in the phloem where they can degrade callose.

(3) Figure 5. The 24-hour time point for callose deposition detection was selected based on established protocols from previous studies. According to Hao et al. (Plant Physiology 2008), callose deposition increased during the ﬁrst 3 days of planthopper infestation and decreased after 4 days. Additionally, Ellinger and Voigt (Ann Bot 2014) demonstrated that callose visualization typically begins 18-24 hours after treatment, making 24 hours an optimal detection time point.

(4) Figure 6, Phloem-specific changes. Similar to Figure 5, while callose deposition was detected in other parts of vascular bundle, significant differences between treatments were mainly observed in phloem cells, indicating that RSV infection specifically affects callose deposition in phloem tissue.

(5) Callose synthase gene expression. We performed RT-qPCR analysis to measure the expression levels of callose synthase genes. The results indicated that OsTLP overexpression did not significantly alter the mRNA levels of these genes, regardless of RSV infection status in SBPH.

Reviewer #2 (Public Review):

There is increasing evidence that viruses manipulate vectors and hosts to facilitate transmission. For arthropods, saliva plays an essential role for successful feeding on a host and consequently for arthropod-borne viruses that are transmitted during arthropod feeding on new hosts. This is so because saliva constitutes the interaction interface between arthropod and host and contains many enzymes and effectors that allow feeding on a compatible host by neutralizing host defenses. Therefore, it is not surprising that viruses change saliva composition or use saliva proteins to provoke altered vector-host interactions that are favorable for virus transmission. However, detailed mechanistic analyses are scarce. Here, Zhao and coworkers study transmission of rice stripe virus (RSV) by the planthopper Laodelphax striatellus. RSV infects plants as well as the vector, accumulates in salivary glands and is injected together with saliva into a new host during vector feeding.

The authors present evidence that a saliva-contained enzyme - carbonic anhydrase (CA) - might facilitate virus infection of rice by interfering with callose deposition, a plant defense response. In vitro pull-down experiments, yeast two hybrid assay and binding affinity assays show convincingly interaction between CA and a plant thaumatin-like protein (TLP) that degrades callose. Similar experiments show that CA and TLP interact with the RSV nuclear capsid protein NT to form a complex. Formation of the CA-TLP complex increases TLP activity by roughly 30% and integration of NT increases TLP activity further. This correlates with lower callose content in RSV-infected plants and higher virus titer. Further, silencing CA in vectors decreases virus titers in infected plants.

(1) Interestingly, aphid CA was found to play a role in plant infection with two non-persistent non-circulative viruses, turnip mosaic virus and cucumber mosaic virus (Guo et al. 2023 doi.org/10.1073/pnas.2222040120), but the proposed mode of action is entirely different.

We appreciate the reviewer’s insightful comment and have carefully examined the cited publication. The study by Guo et al. (2023) elucidates a distinct mechanism for aphid-mediated transmission of non-persistent, non-circulative viruses (turnip mosaic virus and cucumber mosaic virus). In their model, aphid-secreted CA-II in the plant cell apoplast leads to H+ accumulation and localized acidification. This trigger enhanced vesicle trafficking as a plant defense response, inadvertently facilitating virus translocation from the endomembrane system to the apoplast.

In contrast to these pH-dependent mechanisms, our study demonstrated that LssaCA’s biological function in mediating RSV infection is, if not completely, at least partially independent of its enzymatic activity. We performed additional experiments to reveal that OsTLP enzymatic activity is highly pH-dependent and exhibits increased enzymatic activity as pH decreases from 7.5 to 5.0 (Fig. 3H); however, the LssaCA-OsTLP interaction occurring at pH 7.4 significantly enhanced OsTLP enzymatic activity without any change in buffer pH (Fig. 3G). These results demonstrate the crucial importance of LssaCA-OsTLP protein interactions, rather than enzymatic activity alone, in mediating RSV infection.

We have incorporated these new experimental results and added a comprehensive discussion comparing the aphid CA mechanism described by Guo et al. (2023) with our findings in the revised manuscript. Please refer to Figures 3G-H, Lines 220-227 and 350-371 for detailed information.

(2) While this is an interesting work, there are, in my opinion, some weak points. The microinjection experiments result in much lower virus accumulation in rice than infection by vector inoculation, so their interpretation is difficult.

We acknowledge the reviewer's concern regarding the lower virus accumulation observed in microinjection experiments compared to vector-mediated inoculation. We have removed these experiments from the revised manuscript. To address the core question raised by these experiments, we have conducted new experiments that directly demonstrate the importance of LssaCA-OsTLP protein-protein interactions in mediating RSV infection. These results demonstrate the crucial importance of LssaCA-OsTLP protein interactions, rather than enzymatic activity alone, in mediating RSV infection. Additionally, we have incorporated a comprehensive discussion examining carbonic anhydrase activity, pH homeostasis, and viral infection. Please refer to the detailed experimental results and discussion in the sections mentioned in our previous response (Figures 3G-H, Lines 220-227 and 350-371).

(3) Also, the effect of injected recombinant CA protein might fade over time because of degradation or dilution.

We appreciate the reviewer’s insightful comment. This is indeed a valid concern that could affect the interpretation of microinjection results. To address the temporal dynamics of CA protein presence in planta, we conducted time-course experiments to monitor the retention of naturally SBPH-secreted CA proteins in rice plants. Our analysis at 1- and 3- days post-feeding (dpf) revealed that CA protein levels decreased progressively following SBPH feeding, but could also been detected at 3dpf (Fig. 2H). Please refer to Figures 2H and lines 184-193 for detailed information.

(4) The authors claim that enzymatic activity of CA is not required for its proviral activity. However, this is difficult to assess because all CA mutants used for the corresponding experiments possess residual activity.

We appreciate the reviewer’s insightful comment. We constructed a comprehensive CA mutant protein in which all seven residues constituting the enzymatic active center mutated (LssaCAH111D, LssaCAN139H, LssaCAH141D, LssaCAH143D, LssaCAE153H, LssaCAH166D, LssaCAT253E) (Fig. S1B). This LssaCA mutant protein demonstrated complete loss of enzymatic activity (Fig. 1C). However, since we have removed the recombinant CA protein microinjection experiments from the revised manuscript, we lack sufficient direct evidence to definitively demonstrate that CA enzymatic activity is dispensable for its proviral function. To address the core question raised by these experiments, we have conducted new experiments that provide direct evidence for the importance of LssaCA-OsTLP protein-protein interactions in mediating RSV infection. Additionally, we have incorporated a comprehensive discussion examining carbonic anhydrase activity, pH homeostasis, and viral infection. Please refer to the detailed experimental results and discussion in the sections mentioned in our previous response (Figures 3G-H, Lines 220-227 and 350-371).

(5) It remains also unclear whether viral infection deregulates CA expression in planthoppers and TLP expression in plants. However, increased CA and TLP levels could alone contribute to reduced callose deposition.

We have compared LssaCA mRNA levels in RSV-free and RSV-infected L.striatellus salivary glands, which indicated that RSV infection does not significantly affect LssaCA expression (Figure 1J). By using RSV-free and RSV-infected L.striatellus to feed on rice seedlings, we clarified that RSV infection does not affect TLP expression in plants (Figure 5H).

Reviewer #1: (Recommendations For The Authors):

Other comments:

(1) Most data proving viral infection and LssaCA expression were derived from qPCR assays. Western blot data are strongly required to prove the change at the protein level.

We agree that western blot data are required to prove the change at the protein level. In the revised manuscript, we have added western-blotting results (Figures 1F, 1I, 2C, 2J, and S6).

(2) Line 145, data that LssaCA was significantly downregulated should be shown.

Thank you and the data has been added to the revised manuscript. Please refer to Line 165 and Figure 2D.

(3) Lines 159-161, how did authors assure that the dose of recombinant LssCA was closed to the release level of insect feeding, but not was excessive? How did author exclude the possibility of upregulated RSV titer caused by excessive recombinant LssCA?

We appreciate this important concern regarding dosage controls. While microinjection of recombinant proteins typically yields viral infection levels significantly lower than those achieved through natural insect feeding, higher protein concentrations are often required to achieve high viral infection levels. In this experiment, we compared RSV infection levels following microinjection of BSA+RSV versus LssaCA+RSV, with the expectation that any observed upregulation in RSV titer would be specifically attributable to recombinant LssaCA rather than excessive protein dosing. However, given the low RSV infection levels observed with viral microinjection, we have removed their corresponding results from the revised manuscript.

(4) Lines 124-125, recombinantly expressed LssaCA protein should be underlined, but not the LssaCA protein itself.

We have clearly distinguished recombinantly expressed LssaCA from endogenous LssaCA protein throughout the manuscript, ensuring that all references to recombinant proteins are properly labeled as such.

(5) LssaCA expression in salivary glands of viruliferous and nonviruliferous insects is required. LssaCA accumulation in rice plant exposed to viruliferous and nonviruliferous insects is also required.

We have measured LssaCA mRNA levels in salivary glands of viruliferous and nonviruliferous insects (Figure 1J), and protein levels in rice plant exposed to viruliferous and nonviruliferous insects (Figure 1I).

(6) Fig. 4G, the enzymatic activities of OsTLP were too low compared with that in Fig. 4E and Fig. 7E. Why did the enzymatic activities of the same protein show so obvious difference?

We apologize for the error in Fig. 4G. The original data presented relative fold changes between OsTLP+BSA and OsTLP+LssaCA treatment, with OsTLP+BSA normalized to 1.0 and OsTLP+LssaCA values expressed as fold changes relative to this baseline. However, the Y-axis was incorrectly labeled as “β-1,3-glucanase (units mg-1)”, which suggested absolute enzymatic activity values. We have now corrected the figure (revised Figure 3G) to display the actual absolute enzymatic activity values with the appropriate Y-axis label “β-1,3-glucanase (units mg-1)”.

(7) Fig. 7E, was the LssaCA + NP and LssaCA + GST quantified?

Yes, all proteins were quantified, and enzymatic activity values were calculated and expressed as units per milligram of proteins (units mg-1).

Minor comments:

(1) The keywords: In fact, the LssaCA functioned during initial viral infection in plant, but not viral horizontal transmission.

We appreciate the reviewer’s insightful comment. We have revised the manuscript title to “Rice stripe virus utilizes an Laodelphax striatellus salivary carbonic anhydrase to facilitate plant infection by direct molecular interaction” and changed the keyword from “viral horizontal transmission” to “viral infection of plant”.

(2) Fig. 2A, how about testes? Was this data derived from female insects? Fig. 2C, is the saliva collected from nonviruliferous insects? Fig. 2E, what is the control?

We appreciate the reviewer’s insightful comments.

(1) Fig. 2A: The data present mean and SD calculated from three independent experiments, with 5 tissue samples per experiment. Since 3rd instar nymphs were used for feeding experiments in this study, we also used 3rd instar RSV-free nymphs to measure gene expression in guts, salivary glands and fat bodies. R-body represents the remaining body after removing these tissues. Female insects were used to measure gene expression in ovaries, and gene expression in testes was also added. We have added this necessary information to the revised manuscript (please refer to new Figure 1F and Lines 402-403).

(2) Fig. 2C: Yes, saliva was collected from nonviruliferous insects.

(3) Fig. 2E: The control consisted of 100 mM PBS, as described in the experimental section (Lines 643-644): “A blank control consisted of 2 mL of 100 mM PBS (pH 7.0) mixed with 1 mL of 3 mM p-NPA.” In the revised manuscript, we recombinantly expressed LssaCA and its mutant proteins in both sf9 cells and E.coli. Therefore, we have used the mutant proteins as controls to demonstrate specific enzymatic activity. Please refer to Figure 1C, Lines 115-122 and 621-635 for detailed information.

(3) Some figure labeling appeared unprofessional. For example, "a-RSV", "loading" in Fig. 1, "W-saliva", "G-saliva" in Fig. 2, and so on, the related explanations were absent.

We appreciate the reviewer’s insightful comments. We have thoroughly reviewed all figures to ensure professional labels. Specifically, we have:

(1) Used proper protein names to label western blots and clearly explained the antibodies used for protein detection.

(2) Provided comprehensive explanations for all abbreviations used in figures within the corresponding figure legends.

(3) Ensured consistent and clear labeling throughout all figures.

Please refer to the revised Figures 1-3 for these corrections.

(4) Lines 83-84, please cite references on callose preventing viral movement. I do not think the present references were relevant.

We have added a more relevant reference (Yue et al., 2022, Line 82), which revealed that palmitoylated γb promotes virus cell-to-cell movement by interacting with NbREM1 to inhibit callose deposition at plasmodesmata.

(5) The background of transgenic plants of OsTLP OE should be characterized. And the overexpression of OsTLP should be shown. Which generation of OsTLP OE did authors use?

The background of transgenic plants of OsTLP OE and its generation used have been shown in the “Materials and methods” section (Line 782-786) and has been mentioned in the main text (Line 214). T2 lines have been selected for further analysis (Line 789).

(6) Fig. 5A, the blank, which derived from plants without exposure to insect, was absent.

We appreciate the reviewer’s insightful comments. We have added the non- fed control in the revised Figure 5A-C.

(7) Fig. 7A, the nonviruruliferous insects were required to serve as a control.

Immunofluorescence localization of RSV and LssaCA in uninfected L. striatellus salivary glands have been added to the revised manuscript (Figure S2).

(8) The manuscript needs English language edit.

The manuscript has undergone comprehensive English language editing to improve clarity, grammar, and overall readability.

Reviewer #2 (Recommendations For The Authors):

(1) The first experiment compares vector inoculation vs microinjection of RSV in tissue. I am not sure that your claim (saliva factors are necessary for inoculation) holds, because the vector injects RSV directly into the phloem, whereas microinjection is less precise and you cannot control where exactly the virus is deposed. However, virus deposited in other tissues than the phloem might not replicate, and indeed you observe, compared to natural vector inoculation, highly reduced virus titers.

We appreciate the reviewer’s insightful comments. We agree that the comparison between vector inoculation and microinjection involves multiple confounding factors that could affect the experimental results, including salivary components, RSV inoculation titers, and the precision of viral deposition. As the reviewer correctly points out, the differential outcomes could be attributed to these various factors rather than definitively demonstrating the necessity of salivary factors. Therefore, we have removed this comparison from the revised manuscript and instead focused on elucidating the specific mechanisms by which LssaCA facilitates viral infection.

(2) Next the authors show that a carbonic anhydrase (CA) that they previously detected in saliva is functional and secreted into rice. I assume this is done with non-infected insects, but I did not find the information. Silencing the CA reduces virus titers in inoculated plants at 14 dpi, but not in infected planthoppers. At 1 dpi, there is no difference in RSV titer in plants inoculated with CA silenced planthoppers or control hoppers. To see a direct effect of CA in virus infection, purified virus is injected together with a control protein or recombinant CA into plants. At 14 dpi, there is about double as much virus in the CA-injected plants, but compared to authentic SBPH inoculation, titers are 20,000 times lower. Actually, I believe it is not very likely that the recombinant CA is active or present so long after initial injection.

We appreciate the reviewer’s insightful comments.

(1) Our previous study identified the CA proteins from RSV-free insects. We have added this information to the revised manuscript (Line 110).

(2) We acknowledge the reviewer's concern regarding the lower virus accumulation observed in microinjection experiments compared to vector-mediated inoculation. We have removed these experiments from the revised manuscript and instead focused on elucidating the specific mechanisms by which LssaCA facilitates viral infection.

(3) We didn’t intend to suggest that LssaCA proteins presented for 14 days post-injection. We measured viral titers at 14 days post-feeding or post-injection because this is the common measurement time point when viral titers are sufficiently high for reliable detection by RT-qPCR or western blotting. We have clarified this rationale in the revised manuscript (Lines 155-157). To determine the actual persistence of LssaCA in plant tissues, we monitored LssaCA protein levels at 1 and 3 dpf. Western blotting analysis revealed that LssaCA protein levels decreased post-feeding and remained detectable at 3 dpf. These results are presented in Figure 2H and described in detail in Lines 184-193.

(3) Then the authors want to know whether CA activity is required for its proviral action and single amino acid mutants covering the putative active CA site are created. The recombinant mutant proteins have 30-70 % reduced activity, but none of them has zero activity. When microinjected together with RSV into plants, RSV replication is similar as injection with wild type CA. Since no knock-out mutant with zero activity is used, it is difficult to judge whether CA activity is unimportant for viral replication, as claim the authors.

We appreciate the reviewer’s insightful comment. We constructed a comprehensive CA mutant protein in which all seven residues constituting the enzymatic active center mutated (LssaCAH111D, LssaCAN139H, LssaCAH141D, LssaCAH143D, LssaCAE153H, LssaCAH166D, LssaCAT253E) (Fig. S1B). This LssaCA mutant protein demonstrated complete loss of enzymatic activity (Fig. 1C). However, since we have removed the recombinant CA proteins microinjection experiments from the revised manuscript, we lack sufficient direct evidence to definitively demonstrate that CA enzymatic activity is dispensable for its proviral function. To address the core question raised by these experiments, we have conducted new experiments that provide direct evidence for the importance of LssaCA-OsTLP protein-protein interactions in mediating RSV infection. Additionally, we have incorporated a comprehensive discussion examining carbonic anhydrase activity, pH homeostasis, and viral infection. Please refer to the detailed experimental results and discussion in the sections mentioned in our previous response (Figures 3G-H, Lines 220-227 and 350-371).

(4) Next a yeast two hybrid assay reveals interaction with a thaumatin-like rice protein (TLP). It would be nice to know whether you detected other interacting proteins as well. The interaction is confirmed by pulldown and binding affinity assay using recombinant proteins. The kD is in favor of a rather weak interaction between the two proteins.

We have added a list of rice proteins that potentially interact with LssaCA (Table S1) and have measured interactions with additional proteins (unpublished data). Despite the relatively weak binding affinity, the functional significance of the LssaCA-OsTLP interaction in enhancing TLP enzymatic activity is substantial.

(5) Then the glucanase activity of TLP is measured using recombinant TLP-MBP or in vivo expressed TLP. It is not clear to me which TLP is used in Fig. 4G (plant-expressed or bacteria-expressed). If it is plant-expressed TLP, why is its basic activity 10 times lower than in Fig. 4F?

Fig. 4G is the Fig. 3G in the revised manuscript. A E. coli-expressed TLP protein has been used. We apologize for the error in our original Fig. 4G. The original data presented relative fold changes between OsTLP+BSA and OsTLP+LssaCA treatment, with OsTLP+BSA normalized to 1.0 and OsTLP+LssaCA values expressed as fold changes relative to this baseline. However, the Y-axis was incorrectly labeled as “β-1,3-glucanase (units mg-1)”, which suggested absolute enzymatic activity values. We have now corrected the figure to display the actual absolute enzymatic activity values with the appropriate Y-axis label “β-1,3-glucanase (units mg-1)”.

(6) There is also a discrepancy in the construction of the transgenic rice plants: did you use TLP without signal peptide or full length TLP? If you used TLP without signal peptide, you should explain why, because the wild type TLP contains a signal peptide.

We cloned the full-length OsTLP gene including the signal peptide sequence (Line 782 in the revised manuscript).

(7) The authors find that CA increases glucanase activity of TLP. Next the authors test callose deposition by aniline blue staining. Feeding activity of RSV-infected planthoppers induces more callose deposition than does feeding by uninfected insects. In the image (Fig. 5A) I see blue stain all over the cell walls of xylem and phloem cells. Is this what the authors expect? I would have expected rather a patchy pattern of callose deposition on cell walls. Concerning sieve plates, I cannot discern any in the image; they are easier to visualize in longitudinal sections than in transversal section as presented here.

We appreciate the reviewer’s insightful comment.

(1) Callose deposition pattern: While callose deposition was detected in other parts of the vascular bundle, significant differences between treatments were mainly observed in phloem cells, indicating that phloem-specific callose deposition is the primary response to RSV infection and SBPH feeding (Figures 5B and 5J).

(2) Sieve plate visualization: We have examined longitudinal sections to visualize sieve plates, which revealed callose deposition in sieve plates during SBPH feeding and RSV infection (Figure S7).

(3) Quantitative analysis: In addition to aniline blue staining, we quantified callose concentrations using a plant callose ELISA kit to provide more precise measurements (Figure 5A, 5I and S8A).

(4) Gene expression analysis: We utilized RT-qPCR to measure callose synthase expression in both feeding and non-feeding areas, confirming that callose synthesis was induced specifically in feeding regions, leading to localized callose deposition (Figures 5D-H).

These experimental results collectively demonstrate that RSV infection induces enhanced callose synthesis and deposition, with this response occurring primarily in phloem cells, including sieve plates, within feeding sites and their immediate vicinity.

(8) I do not quite understand how you quantified callose deposition (arbitrary areas?) with ImageJ. Please indicate in detail the analysis method.

We have added more detailed information for the methods to quantify callose deposition (Lines 673-678).

(9) More callose content is also observed by a callose ELISA assay of tissue extracts and supported by increased expression of glucanase synthase genes. Did you look whether expression of TLP is changed by feeding activity and RSV infection? Silencing CA in planthoppers increases callose deposition, which is inline with the observation that CA increases TLP activity.

We measured OsTLP expression following feeding by RSV-free or RSV-infected SBPH and found that gene expression was not significantly affected by either insect feeding or RSV infection. These results have been added to the revised manuscript (Lines 275-277 and Figure 5H).

(10) Next, callose is measured after feeding of RSV-infected insects on wild type or TLP-overexpressing rice. Less callose deposition (after 2 days) and more virus (after 14 days) is observed in TLP overexpressors. I am missing a control in this experiment, that is feeding of uninfected insects on wild type or TLP overexpressing rice, where I would expect intermediate callose levels.

We appreciate the reviewer’s insightful comment and fully agree with the prediction. In the revised manuscript, we have constructed ostlp mutant plants and conducted additional experiments to further clarify how callose deposition is regulated by insect feeding, RSV infection, LssaCA levels, and OsTLP expression. Specifically:

(1) Both SBPH feeding and RSV infection induce callose deposition, with RSV-infected insect feeding resulting in significantly higher callose levels compared to RSV-free insect feeding (Fig. 5A-C).

(2) LssaCA enhances OsTLP enzymatic activity, thereby promoting callose degradation (Fig. 5I-K).

(3) OsTLP-overexpressing (OE) plants exhibit lower callose levels than wild-type (WT) plants, while ostlp mutant plants show higher callose levels than WT (Fig. 6A-B).

(4) In ostlp knockout plants, LssaCA no longer affects callose levels, indicating that OsTLP is required for LssaCA-mediated regulation of callose (Fig. 6C-D).

These additional data address the reviewer’s concern and support the conclusion that OsTLP plays a central role in modulating callose levels in response to RSV infection and insect feeding.

(11) Next the authors test for interaction between virions and CA. Immunofluorescence shows that RSV and CA colocalize in salivary glands; in my opinion, there is partial and not complete colocalization (Fig. 7A).

We agree with the reviewer’s observation. CA is primarily produced in the small lobules of the principal salivary glands, while RSV infects nearly all parts of the salivary glands. In regions where RSV and CA colocalize within the principal glands, the CA signal appears sharper than that of RSV, likely due to the relatively higher abundance of CA compared to RSV in these areas. This may explain the partial, rather than complete, colocalization observed in our original Figure 7A. In the revised manuscript, please refer to Figure 1A.

(12) Pulldown experiments with recombinant RSV NP capsid protein and CA confirm interaction, binding affinity assays indicate rather weak interaction between CA and NP. Likewise in pull-down experiments, interaction between NP, CA and TLP is shown. Finally, in vitro activity assays show that activity of preformed TLP-CA complexes can be increased by adding NP; activity of TLP alone is not shown.

We performed two independent experiments to confirm the influence on TLP enzymatic activity by LssaCA or by the LssaCA-RSV NP complex. In the first experiment, we compared the enhancement of TLP activity by LssaCA using TLP alone as a control (Figure 3G). In the second experiment examining the LssaCA-RSV NP complex effect on TLP activity, we used the LssaCA-TLP combination as the baseline control rather than TLP alone (Figure 4B), since we had already established the LssaCA enhancement effect in the previous experiment.

(13) For all microscopic acquisitions, you should indicate the exact acquisition conditions, especially excitation and emission filter settings, kind of camera used and objectives. Use of inadequate filters or of a black & white camera could for example be the reason why you observe a homogeneous cell wall label in the aniline blue staining assays. Counterstaining cell walls with propidium iodide might help distinguish between cell wall and callose label.

Thank you for your insightful suggestions. We have added the detailed information to the revised manuscript (Lines 656-659 and 673-678).

(14) You should provide information whether CA is deregulated in infected planthoppers, as this could also modify its mode of action.\

We have compared LssaCA mRNA levels in RSV-free and RSV-infected L.striatellus salivary glands. The results indicated that RSV infection does not significantly affect LssaCA expression (Figure 1J).

(15) You should show purity of the proteins used for affinity binding measurements.

We have included SDS-PAGE results of purified proteins in the revised manuscript (Figure S3).

(16) L 39: Not all arboviruses are inoculated into the phloem.

Thank you. We have revised this description (Lines 40, 73, 95 and 97).

(17) L 76: Watery saliva is also injected in epidermis and mesophyll cells.

Thank you. We have revised this description (Line 73).

(18) L 79: What do you mean by "avirulent gene"?

Thank you for your valuable comments. We have revised this description as “certain salivary effectors may be recognized by plant resistance proteins to induce effector-triggered immunity”. Please refer to Lines 76-77 for detail.

(19) L 128: Please add delivery method.

Thank you. We have added the delivery methods (Line 134).

(20) L 195: Please explain "MST".

Explained (Line 124). Thank you.

(21) L 203: Please add the plant species overexpressing TLP.

Added (Line 214). Thank you.

(22) L 213: Callose deposition has also a role against phloem-feeding insects.

We appreciate the reviewer’s insight comment. We have added this information to the revised manuscript (Line 252).

(23) L 626: What is a "mutein"?

"mutein" is an abbreviation for mutant proteins. Since the recombinant protein microinjection experiments have been removed from the revised manuscript, the term “mutein” has also been removed. For all other instances, we now use the full term “mutant proteins”.

(24) Fig. 1E: what is "loading"? You should rather show here and elsewhere (or add to supplement) complete protein gels and Western blot membranes and not only bands of interest.

Thank you for your valuable suggestion. Although Figure 1E has been removed from the revised manuscript, we have carefully reviewed all figures to ensure that the term “loading” has been replaced with the specific protein names where appropriate.

(25) Fig. 2C: Please indicate which is the blot and which is the silver stained gel and add mass markers in kDa to the silver stained gel.

Thank you for your suggestion. We have revised figure to include labeled silver-stained gels with indicated molecular weight markers (Figure 1H in the revised manuscript).

Version 2

Abstract

Abstract

Plant viruses typically rely on insect vectors for transmission between plants, with insect salivary proteins playing critical roles in this process. In this study, we demonstrate how Laodelphax striatellus salivary carbonic anhydrase (LssaCA) promotes rice stripe virus (RSV) infection in plants. We discovered that LssaCA directly binds to RSV nucleocapsid protein (NP) in insect salivary glands. This LssaCA-NP complex interacts with a rice thaumatin-like protein (OsTLP) that possess endo-β-1,3-glucanase activity potentially degrading callose. Upon binding, the LssaCA-NP complex significantly enhances OsTLP enzymatic activity. We further clarify that both L. striatellus feeding and RSV infection induce callose deposition. The tripartite LssaCA-NP-OsTLP interaction enhances callose degradation, thereby facilitating RSV infection via its insect vector. This study provides new insights into complex virus-insect-plant tripartite interactions mediated by insect salivary proteins, with broad implications for numerous plant viruses transmitted by insect vectors.

Introduction

During plant virus transmission by piercing-sucking insects, most viruses are inoculated into the plant phloem via the insect’s secreted saliva (Arca & Ribeiro, 2018; Conway et al., 2016; Wu et al., 2022). Thus, insect saliva acts as an interface for the virus-insect-host tripartite interaction and can directly promote viral transmission to, and infection of, the host plants (Sun et al., 2020; Wu et al., 2022). However, despite the importance of insect salivary proteins in this tripartite interaction, there is still much to learn about how these proteins enable successful viral infection.

Previous studies have revealed that there are two ways in which insect saliva facilitates viral infection. One is an indirect approach whereby the saliva modulates the host microenvironment at the feeding site, and saliva effectors work together to allow the arthropod to go unnoticed while it feeds on the host plant (Acevedo et al., 2019; Arca & Ribeiro, 2018; Sun et al., 2020). For example, our recent work indicated that a Laodelphax striatellus mucin protein, LssaMP, enables the formation of the salivary sheath and facilitates the transmission of rice stripe virus (RSV) into the rice phloem (Huo et al., 2022). The study on leafhoppers revealed that the expression of a saliva calcium-binding protein is inhibited by rice gall dwarf virus (RGDV), thus causing an increase of cytosolic Ca2+ levels in rice and triggering callose deposition and H2O2 production. This increases the frequency of insect probing, thereby enhancing viral horizontal transmission into the rice phloem (Wu et al., 2022). The other mechanism by insect saliva to facilitate virus infection is direct regulation, whereby saliva proteins promote virus transmission through specific molecular interactions (Wen et al., 2019). Direct saliva protein-pathogen interactions have been reported in animal pathogens. For example, during transmission of Borrelia burgdorferi by Ixodes scapularis, the saliva protein Salp15 of I. scapularis binds to the bacterial outer surface protein C, which prevents the bacterium from being recognized by the animal immune system. In this way, the saliva protein enables the pathogen to infect the animal host (Ramamoorthi et al., 2005; Schuijt et al., 2008). Although most plant viruses are heavily dependent on insect vectors for plant-to-plant transmission (Gray, 2008), the direct function of insect saliva proteins in mediating virus transmission remains largely uninvestigated.

During sap-feeding, arthropods produce two distinct types of saliva at different stages of the feeding process: gel saliva and watery saliva (Bonaventure, 2012; Lou et al., 2019). The former forms a salivary sheath to provide a smooth path for the stylet penetration (Lou et al., 2019). The latter is mainly secreted into the phloem sieve elements to prevent them from plugging up and suppress plant defense responses. Some salivary components may act as herbivore-associated molecular patterns that can trigger pattern-triggered immunity, and certain salivary effectors may be recognized by plant resistance proteins to induce effector-triggered immunity, etc. (Huang et al., 2017; Ji et al., 2017; Yi et al., 2021).

For phloem-feeding insects, callose deposited on phloem sieve plate and plasmodesmata of sieve elements functions as a defense mechanism by reducing insect feeding and preventing viral movement (Hao et al., 2008; Hipper et al., 2013; Will & Vilcinskas, 2015; Zavaliev et al., 2011; Yue et al., 2022). Callose is a β-(1,3)-D-glucan polysaccharide that is synthesized by callose synthases and degraded by β-(1,3)-glucanases. Plants defend themselves by depositing callose at the sieve plates and plasmodesmata in response to virus infection, whereas viruses counter this defense by activating β-(1,3)-glucanases to degrade callose (Bucher et al., 2001; Hao et al., 2008; Wu et al., 2022; Zavaliev et al., 2011).

RSV is the causative agent of rice stripe disease, a serious disease of rice crops that has occurred repeatedly in China, Japan, and Korea (Xu et al., 2021). RSV is completely dependent on insect vectors for transmission among its host plants, and L. striatellus is the main vector (Xu et al., 2021; Zhao et al., 2018). L. striatellus transmits RSV in a persistent-propagative manner. The virus initially infects the midgut, then disperses from the hemolymph into the salivary glands, and is inoculated into the plant host during L. striatellus feeding (Huo et al., 2022). L. striatellus belongs to the order Hemiptera, whose members mainly feed from sieve tubes through their mouthparts (stylets) that penetrate plant tissues and reach sieve tubes to ingest the phloem sap (Tjallingii, 2006; Van Bel & Will, 2016). RSV is mainly secreted into the rice phloem via the watery saliva (Huo et al., 2022; Wang & Blanc, 2021).

In this study, we identified a molecular interaction among RSV, an L. striatellus saliva protein, and a plant β-1,3-glucanase. The insect saliva protein directly binds to the RSV nucleocapsid protein (NP) and then binds to a rice thaumatin-like protein to activate its β-1,3-glucanase activity. The activation of β-1,3-glucanase helps RSV infection by inhibiting callose deposition in response to viral infection.

Results

L. striatellus salivary carbonic anhydrase directly bind to RSV NP

Insect-secreted watery saliva is inoculated into plant phloem sieve tubes along with RSV (Huo et al., 2022). We identified the salivary proteins that directly interact with RSV to promote viral plant infection. Among the saliva proteins identified in our previous study from RSV-free L. striatellus (Huo et al., 2022), Sap6 (GenBank accession No. RZF48846.1) was found to co-localize with RSV in infected salivary glands, suggesting a potential molecular interaction between them (Fig. 1A and S2).

An L. striatellus salivary carbonic anhydrase (LssaCA) directly binds to RSV NP in salivary glands

A. Immunofluorescence showing co-localization of RSV (red) and LssaCA (green) in L. striatellus salivary glands. Images represent three independent experiments with 30 SBPHs analyzed. Scale bar: 20 μm. psg, principal salivary gland; asg, accessory salivary gland.

B. Schematic of LssaCA protein sequence. SP, signal peptide; Carb_anhydrase domain, conserved eukaryotic-type carbonic anhydrase sequence. Triangles indicate seven predicted catalytically active residues.

C. Esterase activity of recombinant LssaCA protein expressed in sf9 insect cell and E. coli system. In LssaCA mutant protein, all seven predicted catalytically active residues were replaced. Mean and SD were calculated from 2 independent enzymatic assays. ****, P<0.0001.

D. GST pull-down assays demonstrating interaction between LssaCA (GST-LssaCA) and RSV NP (NP-His). GST served as negative control. Anti-GST and anti-His antibodies detected corresponding proteins.

E. MST assay showing binding between LssaCA (LssaCA-His) and RSV NP (GST-NP). GST served as negative control. Error bars represent SD.

F. Measurement of LssaCA tissure-specific expression by RT-qPCR and western blotting. Third-instar RSV-free nymphs were used to measure the gene expression in gut, salivary gland, fatbody, and remaining tissues (R-body) after dissection of the aforementioned tissues. Adult females and adult males were used to measure gene expression in ovaries and testes, respectively. Mean and SD were calculated from three independent experiments with 5 tissue samples each experiment. LssaCA protein was detected using anti-LssaCA antibodies.

G. Western blot analysis of LssaCA protein in RSV-free SBPH salivary glands. dsLssaCA treatment significantly reduces target protein levels compared to controls. LssaCA protein was detected using anti-LssaCA antibodies.

H. Western blot analysis showing LssaCA distribution in watery and gel saliva of RSV-free insects. LssaCA protein was detected using anti-LssaCA antibodies. Total saliva proteins were visualized by silver staining.

I. Western blot detection of LssaCA proteins in rice phloem following L. striatellus feeding. LssaCA protein was detected using anti-LssaCA antibodies.

J. Analyses of LssaCA levels regulated by RSV infection. Third-instar RSV-free and RSV-infected nymphs were used to measure gene expression (RT-qPCR) and protein accumulation (Western blotting). Mean and SD were calculated from three independent experiments with totally 14 tissue samples each experiment. ns, not significant.

All western blot images are representative of three independent experiments.

Sequence homology analysis revealed Sap6 as a carbonic anhydrase-like protein, with a signal peptide located at the N-terminus (amino acids 1-24) (Fig. 1B and S1A). We recombinantly expressed this protein in E.coli or in sf9 cells as a secreted protein. The purified proteins were measured enzymatic activity in hydrolyzing p-nitroethyl acetate. The observed hydrolyzing activity confirmed Sap6 as a carbonic anhydrase (abbreviated as CA, Fig. 1C). Sequence analysis indicated seven residues constituting the enzymatic active center: H111, N139, H141, H143, E153, H166, and T253. We then constructed a mutant with substations of all seven residues (Fig. S1B. H111D, N139H, H141D, H143D, E153H, H166D, and T253E). The mutant protein lost enzymatic activity, indicating the necessity of the enzymatic active center (Fig. 1C). Sap6 was subsequently named L. striatellus salivary carbonic anhydrase (LssaCA).

We performed pull-down and Microscale thermophoresis (MST) assays to confirm the molecular interaction between LssaCA and RSV NP. Pull-down assays revealed a specific interaction between LssaCA and NP (Fig. 1D). MST assays determined that the KD of the interaction between LssaCA and NP was 2.7 ± 2.2 μM, further confirming the molecular interaction (Fig. 1E and S3).

We then performed quantitative reverse-transcriptional PCR (RT-qPCR) to determine the gene expression and western blotting to assess protein distribution. RT-qPCR revealed specific expression of LssaCA in L. striatellus salivary glands (Fig. 1F), while western blotting confirmed the protein’s presence in this tissue (Fig. 1F).

When LssaCA expression was knocked down using LssaCA-specific double-stranded RNA (dsLssaCA) microinjection, LssaCA protein levels in salivary glands decreased dramatically (Fig. 1G). To investigate LssaCA secretion, we allowed RSV-free insects to feed on an artificial diet, collecting watery saliva proteins from the diet solution and gel saliva proteins from the feeding parafilm. Western blotting demonstrated that the LssaCA protein was secreted as a component of watery saliva (Fig. 1H). Furthermore, by measuring LssaCA protein levels in plants after feeding by RSV-free insects, we confirmed that the LssaCA proteins can be secreted into plants (Fig. 1I). We also examined the regulation of LssaCA expression by RSV. RSV infection did not significantly influence LssaCA mRNA levels in the salivary glands; however, the protein levels was significantly higher in RSV-infected salivary glands (Fig. 1J).

In conclusion, these results indicate that when RSV NP is secreted by L. striatellus, a salivary-gland-specifically expressed protein, LssaCA, can directly bind to RSV and be secreted via saliva.

LssaCA enhances RSV infection in plants

To determine the role of LssaCA in RSV infection of rice plants, 36.8 nL dsLssaCA at 1 ng/nL was microinjected into the hemocoel of RSV-infected 3rd-instar nymphs to interfere with gene expression. At 2 days post-microinjection (dpi), LssaCA mRNA levels were measured by RT-qPCR and compared with the dsGFP control. Results showed that dsLssaCA treatment reduced LssaCA mRNA levels by 80% (Fig. 2A). The two groups of insects were then allowed to feed on healthy rice seedlings (5 insects per rice seedling) for 2 days. RSV NP RNA levels (indicative of RSV titers) in the plants were measured at 14 days post-feeding (dpf), when inoculated viruses had infected the plant host, replicated in plant cells, and viral titers were high enough to be detected by RT-qPCR and western blotting. Compared to rice seedlings fed by dsGFP-treated insects, those fed by dsLssaCA-treated insects had significantly reduced RSV titers (Fig. 2B-C), indicating that LssaCA plays an essential role in mediating RSV infection of rice plants.

LssaCA enhances RSV infection in rice plants

A. Gene silencing efficiency determined by RT-qPCR.

B-C. Analysis of RSV infection levels in rice seedlings. Each plant seedling was fed upon by 5 infected insects for 2 days, followed by 14 days of cultivation. RSV levels were determined by RT-qPCR (B) and western blotting (C). RSV NP detected by anti-NP polyclonal antibodies indicated viral titers. Rubisco large subunit (RBCL) served as protein loading control.

D. Gene silencing efficiency determined by RT-qPCR.

E. RT-qPCR analysis of RSV titers in salivary glands. Each dot represents salivary glands from 5 RSV-infected third-instar nymphs.

F. RSV titers in L. striatellus saliva determined by RT-qPCR. Each dot represents one saliva sample from 10 insects fed with artificial diet.

G. RT-qPCR analysis of RSV titers in rice. Plants were fed upon by infected insects for 24 h (Left: schematic of L. striatellus feeding on specific rice stem sites), with titers assayed immediately post-feeding (Right). Each dot represents an individual rice seedling..

H. Western blot analysis of LssaCA in plants fed upon by 50 RSV-free insects for 2 days and continuously cultured for an additional 1 and 3 days. LssaCA protein was detected using anti-LssaCA antibodies. dpf, days post-feeding.

I-J. RSV infection in plants at 3 dpf. Five RSV-infected third-instar L. striatellus nymphs were fed upon healthy rice plants for 2 days, and viral titers in whole plant shoots were measured at 3 dpf by RT-qPCR (I) and western blotting (J). Each dot represents an individual plant. RSV NP detected by anti-NP polyclonal antibodies indicated viral titers. RBCL protein served as a protein loading control.

Mean and SD were calculated from 3 independent microinjection experiments. ns, not significant; *, P<0.05; **, P<0.01; ****, P<0.0001. All western blot images are representative of three independent experiments.

To determine at which step LssaCA affected RSV infection, we analyzed the influence of LssaCA-deficiency on RSV titers during several steps of the RSV transmission and infection process, including virus load in salivary glands and saliva, and initial inoculation levels in planta. First, at 2 days after dsLssaCA treatment of viruliferous insects, when LssaCA was significantly downregulated (Fig. 2D), RT-qPCR was used to detect RSV titer in insect salivary glands. Results showed that LssaCA deficiency did not reduce RSV titers in this tissue (Fig. 2E). Then, LssaCA-deficient insects were allowed to feed on an artificial diet, and their saliva was collected for RSV level analysis. Results revealed that LssaCA-deficiency did not reduce RSV titer in the secreted saliva (Fig. 2F). Finally, a rice plant was fed with 10 viruliferous insects for 24 h, with insects limited to a specific area, and RSV titers were measured immediately after feeding. We found that LssaCA-deficiency did not affect RSV initial inoculation levels in planta (Fig. 2G). We also monitored the influence of LssaCA deficiency on the insect feeding behaviors by using Electrical penetration graph (EPG). The EPG waveforms can be classified into three different periods: nonpenetration (NP), penetration (N1, N2, N3) and phloem feeding (N4: N4-a, N4-b) periods. The NP waveform indicates a lack of feeding, the N waveform indicates penetration of mesophyll cells to seek the phloem and the N4 waveform indicates digestion of the sap from the phloem. Results indicated that LssaCA deficiency did not affect the insect feeding behaviors (Fig. S4). These results collectly suggest that LssaCA promotes RSV infection through a mechanism occurring not in insects or during early stages of viral entry in plants, but in planta after viral inoculation.

LssaCA protein levels were measured after insect feeding. Plants were fed on by 50 RSV-free insects for 2 days, and then LssaCA protein levels were measured at 1 and 3 dpf by western blotting. We found that LssaCA was retained for at least 3 days, and protein levels decreased rapidly from 1 dpf to 3 dpf (Fig. 2H). We then measured RSV infection levels at 3 dpf to assess the in-planta influence after viral inoculation. LssaCA expression was knocked down using dsLssaCA in RSV-infected 3rd-instar nymphs, and 5 insects were allowed to feed on healthy rice for 2 days. We measured RSV titers in the whole plant shoot at 3 dpf. We found significantly lower RSV titers in the dsLssaCA-treatment group at 3 dpf (Fig. 2I-J). These results further confirmed the enhancement of RSV’s plant infection process by LssaCA.

LssaCA interacts with rice thaumatin-like protein to increase its endo-β-1,3-glucanase activity and facilitate RSV infection in planta

To investigate the mechanism by which LssaCA facilitates RSV infection in plants, we performed yeast two-hybrid screening to identify rice proteins that interact with LssaCA. A rice thaumatin-like protein (XP_025879846.1), designated as Oryza sativa thaumatin-like protein (OsTLP) (Fig. 3A-B and S5, Table S1), was identified. The open reading frame of OsTLP encodes a polypeptide of 327 amino acids with a thaumatin family (THM) domain from residues 27 to 267. A BLAST algorithm-based analysis revealed sequence identities to TLP 1b. The N-terminal 24 amino acids constitute a secretion signal peptide, suggesting OsTLP is a secreted protein. Pull-down and MST assays were performed to confirm the molecular interaction between OsTLP and LssaCA. Pull-down assays showed that OsTLP and LssaCA have a specific interaction (Fig. 3C). The MST assay revealed a KD of 1.3 ± 2.2 μM for the OsTLP-LssaCA interaction, further confirming the molecular interaction (Fig. 3D).

LssaCA interacts with rice thaumatin-like protein (OsTLP) to enhance its endo-β-1,3-glucanase activity

A. Schematic diagram of OsTLP protein sequence. SP, signal peptide; THM, conserved thaumatin family protein sequence. Glyco_hydro_64 domain, glycoside hydrolases of family 64.

B. Yeast two-hybrid assay showing interaction between OsTLP and LssaCA. SD, synthetically defined medium; Leu, Leucine; Trp, Tryptophan; His, Histidine; 3’AT, 3-amino-1,2,4-triazole; AD, transcription activation domain; BD, DNA-binding domain.

C. Pull-down assays showing interactions between LssaCA (LssaCA-His) and OsTLP (MBP-OsTLP). MBP served as a negative control. Anti-His and anti-MBP polyclonal antibodies were used for protein detection. Western blot images are representative of three independent experiments.

D. MST assay showing molecular interactions between LssaCA (LssaCA-His) and OsTLP (MBP-OsTLP). MBP served as a negative control. Error bars represent SD.

E. Endo-β-1,3-glucanase activity of purified recombinant OsTLP protein expressed as MBP-fusion protein. MBP was used for normalization.

F. Endo-β-1,3-glucanase activity in OsTLP-overexpressing transgenic plants (OsTLP OE) or wild-type plants (WT). Plant total soluble proteins were used for enzymatic activity assays. Units mg-1 means units per mg of plant total soluble proteins.

G. Regulation of OsTLP enzymatic activity by LssaCA binding.

H. OsTLP enzymatic activity measured at pH 5.0, 5.5, 6.0, 6.5, 7.0 and 7.5 to assess pH dependence.

I. RT-qPCR analysis of RSV infection in OsTLP OE and WT plants. RSV titers (NP copy number) were measured at 14 dpf. ns, not significant; *, P<0.05; ***, P<0.001; ****, P<0.0001.

Previous studies have shown that TLP orthologs in various plants, including apples, cherries, tomatoes, and barley, possess endo-β-1,3-glucanase activity that degrades callose (de Jesus-Pires et al., 2020). To verify the endo-β-1,3-glucanase activity of OsTLP, we conducted enzymatic assays using recombinantly expressed OsTLP protein. The purified protein exhibited an average activity of 60 units/mg (Fig. 3E). To examine OsTLP activity in planta, we constructed transgenic plants overexpressing OsTLP (OsTLP OE) by using the Japonica rice cultivar Nipponbare (Fig. S6A-B). These transgenic plant showed significantly increased β-1,3-glucanase activity compared to wild-type controls (Fig. 3F).

To determine whether the interaction between LssaCA and OsTLP affects enzymatic activity, we pre-incubated OsTLP with either LssaCA or BSA (control) for 2 h before measuring activity. Pre-incubation with LssaCA significantly enhanced OsTLP activity compared to the BSA control (Fig. 3G). Since CA family proteins catalyze the conversion of CO2 and H2O to HCO₃⁻ and H⁺, potentially lowering pH, we examined the pH dependence of OsTLP activity. Activity assays conducted across a pH range of 5.0-7.5 revealed that OsTLP activity increased at lower pH values (Fig. 3H). However, pH monitoring during OsTLP-LssaCA interaction experiments showed no change from the initial pH of 7.4, indicating that the enhanced enzymatic activity was not due to pH changes but rather to the direct interaction between OsTLP and LssaCA.

Moreover, we measured RSV infection levels in plants with increased OsTLP activity. Viruliferous L. striatellus fed on either OsTLP OE plants or wild-type plants for 2 days, and RSV titers were measured at 14 dpf. RT-qPCR analyses indicated that compared with wild-type plants, plants overexpressing OsTLP had significantly higher RSV titers (Fig. 3I), indicating that OsTLP facilitates RSV infection in planta.

Collectively, these results suggest that LssaCA may promote RSV infection by enhancing OsTLP’s callose-degrading activity.

RSV NP-LssaCA-OsTLP tripartite interaction further enhances OsTLP enzymatic activity

Given that LssaCA interacts with both RSV NP and OsTLP, we investigated the possibility of a tripartite interaction. Through pull-down assays, we discovered that when LssaCA was pre-incubated with NP to form an NP-LssaCA complex, this complex retained the ability to bind to OsTLP (Fig. 4A). This result demonstrates a tripartite interaction among these three proteins.

RSV NP-LssaCA-OsTLP tripartite interaction further enhances OsTLP enzymatic activity

A. Pull-down assays showing interactions between OsTLP (MBP-OsTLP) and LssaCA-NP complex. LssaCA was expressed with a His tag, and RSV NP was expressed as a GST-fusion protein. LssaCA and NP were pre-incubated before co-incubation with OsTLP. Anti-His, anti-GST, and anti-MBP antibodies were used to detect corresponding proteins.

B. Endo-β-1,3-glucanase activity of OsTLP after co-incubation with LssaCA (LssaCA + GST) and LssaCA-NP complex (LssaCA + NP). LssaCA was pre-incubated with either GST or NP before addition to OsTLP. ****, P<0.0001.

We then explored how the tripartite interaction affects OsTLP enzymatic activity. OsTLP was pre-incubated with either the LssaCA-NP complex or LssaCA alone as a control. After 2 h of incubation, enzyme activity was measured. The results showed that pre-incubation with the LssaCA-NP complex significantly increased OsTLP activity compared to the LssaCA control (Fig. 4B). This indicates that stronger activation of β-1,3-glucanase activity is achieved through the NP-LssaCA-OsTLP tripartite interaction.

LssaCA inhibits insect-feeding and RSV-infection induced callose deposition in rice plants

Callose deposition on sieve plates and plasmodesmata serves as an critical defense mechanism in plants, reducing insect feeding, viral spread, and infection severity (Li et al., 2012; Zavaliev et al., 2011). Given that the LssaCA or LssaCA-NP complex enhances OsTLP’s β-1,3-glucanase activity and potentially promotes callose degradation, we examined callose regulation during feeding by both RSV-free and RSV-infected L. striatellus. To assess feeding-induced callose deposition, we exposed rice seedlings to either RSV-free or RSV-infected L. striatellus for 24 h and analyzed callose levels around the feeding sites. ELISA assays showed that while feeding by RSV-free insects increased callose concentration, feeding by RSV-infected insects triggered an even greater increase (Fig. 5A). Consistent with these findings, microscopic analysis revealed enhanced callose deposition in phloem cells of plants fed upon by RSV-infected L. striatellus compared to those fed by RSV-free insects (Fig. 5B-C and S7). To determine whether this response was localized, we measured callose levels in tissues distant from the feeding sites. No significant differences in callose deposition were observed between control plants and the non-feeding areas (Fig. S8A), indicating that callose accumulation was confined to feeding sites and their immediate vicinity.

LssaCA inhibits L. striatellus-transmitted RSV-induced callose deposition in rice plants

A. Callose concentration quantified by ELISA in leaves of rice plants fed on by RSV-free or RSV-infected L. striatellus. Non-fed plants served as the control (CK).

B. Bright blue fluorescence of cross-sections showing callose deposition at feeding sites. Samples were prepared from the leaf phloem of plants that were not fed or were fed on by RSV-free or RSV-infected L. striatellus. Thin sections were stained with 0.1% aniline blue at 24 h post-feeding and examined under a fluorescence microscope. xy, xylem; ph, phloem. Scale bars: 10 μm.

C. Average fluorescence intensity of callose deposition measured using ImageJ. Eight to ten random sites per sample were selected for intensity quantification.

D–H. Transcript levels of callose synthase genes (gls3, gls4, gls5, gls10) and OsTLP determined by RT-qPCR. Insects were allowed to feed on rice plants for 24 h prior to RNA extraction.

I. Callose concentration quantified by ELISA in leaves of rice plants fed on by dsGFP- or dsLssaCA-treated RSV-infected L. striatellus.

J. Bright blue fluorescence of cross-sections showing callose deposition at feeding sites. Samples prepared from the leaf phloem of plants fed on by dsGFP- or dsLssaCA-treated RSV-infected L. striatellus. Scale bars: 20 μm.

K. Average fluorescence intensity of callose deposition measured using ImageJ. Eight to ten random sites per sample were selected for intensity quantification.

Mean and SD were calculated from three independent experiments. ns, not significant; *, p < 0.05; **, p < 0.01; ***, p < 0.001; ****, p < 0.0001.

To evaluate the molecular mechanisms underlying the induced callose deposition, we analyzed transcript levels of callose synthase genes using RT-qPCR. In agreement with the callose deposition patterns, feeding by RSV-free insects evaluated the expression of callose synthase genes, including gsl3, gsl4, gsl5, and gsl10, while RSV-infection further enhanced their upregulation (Fig. 5D–G and S8B-E). These results further support that both insect feeding and RSV infection promote callose deposition. Notably, the transcriptional levels of these genes remained unchanged in non-feeding areas (Fig. S8B-E). We also measured the OsTLP gene expression and found that neither SBPH feeding nor RSV infection significantly induced OsTLP expression (Fig. 5H).

To investigate whether LssaCA influences callose deposition induced by insect-transmitted RSV, we downregulated LssaCA in RSV-infected L. striatellus using dsRNA microinjection. At 2 dpi, when LssaCA expression was significantly reduced, we allowed the insects to feed on healthy plants for 24 h before measuring callose deposition. Our results showed that plants fed upon by LssaCA-deficient insects accumulated more callose compared to those fed upon by control insects (dsGFP treatment) (Fig. 5I-K). These results indicate that LssaCA inhibits callose deposition.

In conclusion, these results demonstrate that both L. striatellus feeding and RSV infection induce callose deposition, which can be suppressed by LssaCA.

LssaCA mediates in planta callose degradation by enhancing OsTLP enzymatic activity

To confirm whether OsTLP degrades the deposited callose induced by insect-transmitted RSV in planta, we constructed ostlp mutant plants (Fig. S6C). Both ELISA assays and microscopic analysis showed that compared to wild-type plants, ostlp plants exhibited significantly higher callose levels (Fig. 6A-B).

LssaCA mediates in planta callose degradation by enhancing OsTLP enzymatic activity

A. Callose concentration quantified by ELISA in leaves of wild-type (WT), OsTLP overexpression (OsTLP OE), and OsTLP mutant (ostlp) plants.

B. Bright blue fluorescence of cross-sections showing callose deposition in leaf phloem. Plants were not fed on by insects. Scale bars: 10 μm. Average fluorescence intensity of callose deposition was measured using ImageJ. Eight to ten random sites per sample were selected for intensity quantification. xy, xylem; ph, phloem.

C. Callose concentration quantified by ELISA in leaves of ostlp plants fed on by dsGFP- or dsLssaCA-treated RSV-infected L. striatellus.

D. Bright blue fluorescence of cross-sections showing callose deposition at feeding sites. Samples were prepared from the leaf phloem of ostlp plants fed on by dsGFP- or dsLssaCA-treated RSV-infected L. striatellus. Scale bars: 10 μm. Average fluorescence intensity was quantified using ImageJ. Eight to ten random sites per sample were selected for intensity quantification.

Mean and SD were calculated from three independent experiments. ns, not significant; *, p < 0.05. ****, p < 0.0001.

Given that LssaCA mediates callose degradation by enhancing OsTLP enzymatic activity, we hypothesized that when OsTLP is knocked out in plant, LssaCA would not affect callose levels induced by RSV infection. To test this, we allowed dsLssaCA-treated viruliferous L. striatellus or dsGFP-treated control insects to feed on ostlp seedlings. ELISA assays indicated that LssaCA deficiency did not significantly change callose accumulation (Fig. 6C). Microscopic analysis confirmed higher callose deposition in phloem cells of ostlp seedlings, which remained unaffected by LssaCA deficiency (Fig. 6D). These results indicate that OsTLP is required in planta for LssaCA-mediated effects on callose deposition.

Discussion

Insects are known to be vectors of many plant viruses, and some of these plant viruses rely entirely on insect vectors for their plant-to-plant transmission. Insect salivary proteins have been reported to facilitate viral infection of plants through various mechanisms, such as breaking down plant cell walls, inhibiting plant immune defenses and callose deposition, directly binding to viruses to prevent their detection by plant immune systems, and more. In this study, we found that RSV infection of rice plants triggers the deposition of callose. However, this process is reversed by a tripartite interaction involving the central component of an insect salivary protein LssaCA, which subsequently binds to RSV in insect salivary glands and then OsTLP in the plants. The activity of OsTLP’s β-1,3-glucanase is significantly increased when this tripartite interaction occurs; leading to the degradation of callose and thus allowing for efficient RSV infection (Fig. 7).

Proposed model of RSV NP-LssaCA-OsTLP tripartite interaction facilitating callose degradation and promoting RSV infection.

In L. striatellus salivary glands (SG), RSV (red star) binds to LssaCA proteins (blue triangle) via the nucleocapsid protein (NP). Upon delivery into the rice phloem, RSV infection triggers callose synthesis, leading to callose deposition that restricts viral spread. However, LssaCA binds to OsTLP (yellow scissors), enhancing its β-1,3-glucanase activity and promoting callose degradation. The RSV NP-LssaCA complex formed in the insect’s SG can still interact with OsTLP in planta, resulting in significantly higher OsTLP β-1,3-glucanase activity. This enhanced enzymatic activity efficiently degrades callose, thereby facilitating RSV infection.

Callose is a complex polysaccharide, and its deposition in plant phloem plays an important role in defending against insect attacks and viral spread (Zavaliev et al., 2011). When attacked by piercing insects or viruses, plants activate callose deposition on the sieve plates to occlude the flow of phloem sap, thus discouraging feeding and reducing viral spread. This accumulation of callose in plant phloem can be switched on or off through the action of callose synthases and hydrolases (Li et al., 2012; Shangguan et al., 2018). Studies have shown that expression levels of these glucanase enzymes differ between resistant and susceptible plant species. This suggests that some plants may be more successful at resisting infection due to high levels of glucanase enzymes in their tissues, which allow them to more quickly and effectively form callose barriers against insects and viruses (Hao et al., 2008). Some insects have evolved β-1,3-glucanases enzymes that can degrade the deposited callose, thus enabling them to bypass the plant’s defense mechanisms and continue feeding (Bucher et al., 2001; Hao et al., 2008). In this study, we demonstrate a novel mechanism by which an insect saliva protein binds to a plant β-1,3-glucanases protein, activating the enzymatic activity and promoting callose degradation.

In this study, we found that plants fed on by viruliferous insects exhibited higher callose deposition than those fed on by uninfected insects (Fig. 5A–C), suggesting that viral infection via insect saliva may constitute a dual stress to the plant, inducing enhanced callose deposition that must be overcome for successful insect feeding and viral spread. To investigate the underlying mechanism, we examined the tripartite interaction between LssaCA-NP-OsTLP. We found that this interaction induced significantly higher β-1,3-glucanase activity than the LssaCA-OsTLP interaction alone, suggesting that SBPH and RSV have evolved mechanisms to enhance β-1,3-glucanase activity in order to bypass the callose-mediated defense deposition and facilitate successful viral spread.

Carbonic anhydrase catalyzes the reversible hydration of carbon dioxide according to the reaction CO₂ + H₂O ⇌ HCO₃⁻ + H⁺. This enzymatic activity is essential for maintaining pH homeostasis in biological systems by regulating the balance between CO₂, bicarbonate, and protons. The importance of carbonic anhydrases in pH regulation is exemplified in mosquito larvae, which utilize an extremely alkaline digestive strategy (pH 10.5-11) that is relatively rare in nature, with carbonic anhydrases serving as the central enzymatic system for generating and maintaining bicarbonate/carbonate buffering in the midgut lumen (Linser et al., 2009). Similarly, carbonic anhydrases in arthropod saliva can significantly impact host-pathogen interactions through pH modulation. A recent study elucidated how aphid salivary carbonic anhydrase II (CA-II) facilitates virus transmission, with the enzyme showing significantly higher expression levels in winged morphs compared to wingless counterparts. When secreted into the plant apoplastic space, CA-II catalyzes H⁺ accumulation, resulting in apoplastic acidification from 5.5 to 5.0. This pH reduction enhances polygalacturonase activity, thereby promoting the degradation of demethylesterified homogalacturonan components. Consequently, plants respond by accelerating vesicle trafficking to transport pectin for cell wall reinforcement, a process that simultaneously facilitates viral translocation from the endomembrane system to the apoplastic compartment (Guo et al., 2023). In contrast to these pH-dependent mechanisms, our study demonstrated that LssaCA’s biological function in mediating RSV infection is, if not completely, at least partially independent of its enzymatic activity. We found that although OsTLP enzymatic activity is highly pH-dependent and exhibits increased enzymatic activity as pH decreases from 7.5 to 5.0 (Fig. 3H), the LssaCA-OsTLP interaction occurring at pH 7.4 significantly enhanced OsTLP enzymatic activity without any change in buffer pH (Fig. 3G). These results demonstrate the crucial importance of LssaCA-OsTLP protein interactions, rather than enzymatic activity alone, in mediating RSV infection. While the exact role of LssaCA enzymatic activity in aiding RSV infection remains unclear, the local low-pH plant environment would enhance OsTLP enzymatic activity, suggesting potential synergistic effects between enzymatic and non-enzymatic functions.

In conclusion, this study elucidated a tripartite interaction among a plant virus, an insect saliva protein, and a host phloem β-1,3-glucanase. This interaction inhibited the plant defense of callose deposition and enabled viral phloem transmission. This study provides new insights into the tripartite virus-insect vector-plant interaction, which is relevant to many agriculturally important plant arboviruses whose transmission is facilitated via insect saliva proteins.

Materials and methods

Viruses, insects, host plants, and antibodies

The RSV-free and RSV-infected L. striatellus individuals used in this study were originally captured in Jiangsu Province, China, and were maintained in our laboratory. All plants used to rear L. striatellus were grown in a growth incubator (14.5 cm height, 3.2 cm radius) at 25℃ under a 16-h light/8-h dark photoperiod. Insects were transferred to fresh seedlings every 15 days to ensure adequate nutrition and insect activity.

RSV ribonucleoproteins (RNPs) were purified from RSV-infected rice plants. These purified RNPs were subsequently used as the antigen to immunize rabbits for the production of an RSV-specific antibody. The LssaCA polyclonal antibody was produced using recombinantly expressed LssaCA protein (with a His-tag at the C-terminus) as the antigen. Anti-His, anti-GST, and anti-MBP antibodies were purchased from ABclonal.

Tissue collection

Insect dissection and tissue collection were performed using previously described procedures (Huo et al., 2018). In brief, insects were anesthetized at 4°C for 10 min, and the forelegs were severed at the coxa-trochanter joint using forceps. The hemolymph was expelled and drawn to the tip of clean forceps. The insects were then dissected from the abdomen in pre-chilled PBS buffer. The fat body was collected with forceps, aided by the tension of the liquid. Tissues, including the guts, salivary glands, and ovaries, were washed three times in PBS to remove contaminants from the hemolymph.

Bioinformatic analysis

The signal peptide and transmembrane helix were predicted using tools at the SignalP 5.0 server (https://services.healthtech.dtu.dk/service.php?SignalP-5.0) and the TMHMM-2.0 server (https://services.healthtech.dtu.dk/service.php?TMHMM-2.0). Homology searches of the protein sequence of LssaCA and OsTLP were conducted using the BLAST algorithm at the National Center for Biotechnology Information (http://blast.ncbi.nlm.nih.gov/Blast.cgi). Conserved domains were predicted with SMART (http://smart.embl-heidelberg.de/).

Collection of watery and gel saliva

Saliva was collected as described previously (Huo et al., 2022). Watery saliva was obtained by centrifuging the collected artificial diet at 10,000 ×g for 10 min to remove the solid gel. To collect gel saliva proteins, the membrane was washed twice with PBS. Sheathes adhering to the parafilm were denatured and solubilized in a solution consisting of 8 M urea, 4% v/v CHAPS, 0.1% w/v sodium dodecyl sulfate (SDS), and 2% v/v DTT. The parafilm was incubated on an orbital shaker at room temperature for 45 min. Then, the denatured gel saliva was concentrated to a final volume of 100 µL. Saliva samples were collected from 2,000 insects for silver staining and western blotting analyses.

Measurement of carbonic anhydrase activity

DNA fragments encoding full-length LssaCA and a LssaCA mutant (H111D, N139H, H141D, H143D, E153H, H166D, and T253E) were amplified by PCR. The resulting products were purified and cloned into the vector pET-30a(+) and pFastBac1 vectors to generate His-tagged recombinant proteins. The pET-30a(+)-LssaCA and pET-30a(+)-LssaCA(mutant) plasmids were transformed into E. coli for protein expression. In parallel, the pFastBac1-LssaCA and the pFastBac1-LssaCA(mutant) plasmids were transformed into E. coli DH10Bac for transposition into the bacmid. Recombinant bacmids were then transfected into Spodoptera frugiperda (Sf9) cells using Cellfectin II Reagent (Invitrogen, USA). Recombinant proteins expressed in both systems were purified using Ni Sepharose resin (GE Healthcare, Uppsala, Sweden).

Carbonic anhydrase activity was assessed by measuring esterase activity using p-nitroethyl acetate (p-NPA) as the substrate. LssaCA and its mutant were expressed in E. coli and Sf9 cells with C-terminal His-tag, and the purified proteins were diluted in PBS buffer (pH 7.0) to a final concentration of 1 mg/mL. Reaction conditions followed the method described by Verpoorte et al. (1967), with minor modifications. Briefly, 1 mL of freshly prepared 3 mM p-NPA (dissolved in acetone) was added to 1.9 mL of 100 mM PBS (pH 7.0), followed by 0.1 mL of LssaCA enzyme solution (0.5 mg/mL in 100 mM PBS). Absorbance at 384 nm was measured at the start of the reaction and after 3 minutes. One unit (U) of enzyme activity was defined as the amount of enzyme required to hydrolyze 1 μmol of p-NPA per minute at pH 7.0 and 25 °C. A blank control consisted of 2 mL of 100 mM PBS (pH 7.0) mixed with 1 mL of 3 mM p-NPA.

Immunofluorescence assays

Immunofluorescence assays (IFA) were performed to determine the localization of RSV and LssaCA in L. striatellus salivary glands. The dissected salivary gland samples were fixed in 4% v/v paraformaldehyde at 4°C overnight and then permeabilized in 2% v/v Triton X-100 for 2 h. The samples were incubated with rabbit anti-RSV serum and mouse anti-LssaCA serum (1:250 dilution in PBST/FBS: PBS containing 2% v/v Triton X-100 and 2% fetal bovine serum) for 1 h at room temperature. After washing twice with PBST, the samples were incubated with 1:200 diluted Alexa Fluor 488-conjugated goat anti-mouse antibody and Alexa Fluor 594-conjugated goat anti-rabbit antibody (Invitrogen, USA) for 1 h. After washing three times with PBST, the samples were examined under a Leica TCS SP8 microscope (Leica, Wetzlar, Germany) with the following settings: 1) Alexa Fluor 488 channel: excitation at 488 nm using an argon laser; emission detected with a 493-553 nm bandpass filter. Alexa Fluor 594 channel: excitation at 594 nm using a HeNe laser; excitation detected with a 600-650 nm bandpass filter.

Evaluation of callose deposition in rice plants

Five insects were allowed to feed for 24 h on a 2 × 2 cm leaf area of 10-d-old rice plants. After removal of the insects, leaf tissues were collected from both the feeding site and an adjacent undamaged area located 1 cm away (non-fed site). Samples were immersed in 5 mL alcoholic lactophenol solution (1 volume of phenol:glycerol:lactic acid:water in a 1:1:1:1 ratio, mixed with 2 volumes of ethanol), evacuated with a syringe for 15 min, and then incubated at 65℃ for 30 min until the chlorophyll was completely cleared. Cleared leaf samples were rinsed with 50% v/v ethanol, then with water, and then stained in 150 mM K2HPO4 (pH 9.5) containing 0.01% w/v aniline blue for 30 min in the dark. Stained leaf samples were sectioned and mounted in 50% v/v glycerol on slides. The slides were examined under a Zeiss observer Z1 immunofluorescence microscope (Carl Zeiss MicroImaging GmbH, Gottingen, Germany) with excitation at 405 nm. Fluorescence intensity was quantified using ImageJ software. For ImageJ analysis, images were converted to 8-bit grayscale and adjusted for brightness. Phloem-specific regions of interest (ROIs) were outlined using the freehand selection tools. A standardized fluorescence threshold (set at >2× the mean background intensity) was applied to isolate callose signals. Integrated density and area values were used to calculate normalized fluorescence intensity. Identical exposure times were maintained across all independent experiments.

Callose concentrations were quantified using a plant callose ELISA kit (Catalog No. JL46407; Jonln, Shanghai, China), following the manufacturer’s instruction. Leaf tissues from feeding and non-fed sites were homogenized in PBS buffer (pH 7.4), and the homogenate was centrifuged at 500–1000 x g for 20 min. The supernatant was collected, and absorbance at 450 nm was measured. Callose concentrations were calculated using a standard curve.

RNA interference

We conducted RNAi analyses to determine the function of LssaCA in mediating RSV infection and regulating callose deposition. The LssaCA-specific gene fragment was PCR-amplified with the primer pair LssaCA-T7F/LssaCA-T7R (Table S1). Then, dsRNA was synthesized using a commercial kit (T7 RiboMAX Express RNAi System, Promega, Madison, WI, USA) and purified by phenol: chloroform extraction and isopropanol precipitation. Finally, 36.8 nL dsRNA at 1 ng/nL was delivered into the insect hemocoel for gene silencing. Microinjection was performed using a glass needle through a Nanoliter 2000 (World Precision Instruments, Sarasota, FL, USA). The negative control, GFP dsRNA, was synthesized and microinjected following the same protocol.

RT-qPCR

Gene transcript levels were determined by RT-qPCR. The RNA was extracted from L. striatellus tissues using a RNeasy Mini Kit (Qiagen, Hilden, Germany), and from plant leaves using TRIzol. Reverse transcriptional PCR (iScript cDNA Synthesis Kit, Bio-Rad, Hercules, CA, USA) and SYBR-Green-based qPCR (SYBR Green Supermix, Bio-Rad) was performed according to the manufacturer’s protocols. The primer pair used to amplify LssaCA was LssaCA-q-F/LssaCA-q-R (Table S1). Viral RNA copies were measured using the primers NP-q-F/NP-q-R (Table S1). L. striatellus translation elongation factor 2 (EF2) and rice actin were amplified as internal controls to ensure equal loading of cDNA isolated from different samples. The primer pairs EF2-q-F/EF2-q-R and actin-q-F/actin-q-R (Table S1) were used to amplify EF2 and actin, respectively. Water was used as a negative control.

Yeast two-hybrid assays

Yeast two-hybrid screening was performed to identify rice proteins interacting with LssaCA. A high-complexity rice cDNA library was fused to Gal4 AD and transformed into the yeast strain Y187. The LssaCA gene lacking the signal-peptide sequence (LssaCAsp-) was fused to Gal4 DNA-BD and transformed into the yeast strain Y2HGold. The two yeast strains were co-cultured overnight to allow mating to create diploids. After library screening, positive clones were selected on quadruple dropout medium: SD/–Ade/–His/–Leu/–Trp supplemented with X-a-Gal and aureobasidin A (QDO/X/A), and prey plasmids were isolated for sequencing. To further confirm the interaction between LssaCA and OsTLP, LssaCAsp- was cloned into the prey vector pDB-Leu, and the OsTLP gene lacking the signal-peptide sequence (OsTLPsp-) was cloned into the bait vector pPC86. The two plasmids were co-transformed into the yeast stain Mav203 and the transformants were selected on SD/–His–Leu–Trp + 20 mM 3-AT medium.

Pull-down assays

Pull-down assays were performed to determine the molecular interactions between LssaCA and RSV NP, LssaCA and OsTLP, and OsTLP and the LssaCA-NP complex. The interaction between LssaCA and RSV NP was determined using GST pull-down analyses as described previously (Huo et al., 2014). Briefly, LssaCAsp- was cloned into the pGEX-3x vector and LssaCA was expressed as a GST-fusion protein (GST-LssaCA). The NP gene was cloned into the pET-30a (+) expression vector and the protein was expressed with an N-terminal His-tag (NP-His). The GST-LssaCA construct was expressed in, and purified from, E. coli stain BL21. The GST-LssaCA protein or the GST control protein was bound to Glutathione Sepharose beads (Cytiva, Marlborough, MA, USA), then NP-His protein was added to the beads. The bead-bound proteins were separated by SDS-PAGE and detected by western blotting with antibodies against GST or His-tag.

Next, MBP pull-down analyses were performed to determine the interaction between OsTLP and LssaCA. The OsTLPsp- sequence was cloned into pMAL-p2X to produce the MBP-fusion protein (MBP-OsTLP). The MBP-OsTLP protein or the MBP control protein was bound to amylose resin (New England Biolabs, Beverley,MA, USA), and then the LssaCA-His protein was added to the beads. To determine the interaction between OsTLP and the LssaCA-NP complex, the LssaCA-His and GST-NP proteins were preincubated before being added to the beads. The bead-bound proteins were separated by SDS-PAGE and detected by western blotting with the MBP antibody (Cell Signaling Technology, Danvers, MA, USA), His-tag antibody (Tiangen, Beijing, China) and GST antibody (Tiangen).

Microscale thermophoresis (MST) assays

We conducted MST assays to detect the interactions between LssaCA and NP, and between LssaCA and OsTLP. First, 10 μM purified LssaCAsp- protein (His-fusion protein) was labelled with a Monolith NT Protein Labeling Kit RED-NHS (Nano Temper Technologies GMBH, München, Germany) using red fluorescent dye NT-647 N-hydroxysuccinimide (amine-reactive), according to the manufacturer’s instructions. The binding assays were performed on a Monolith NT.115 Microscale Thermophoresis instrument (Nano Temper Technologies GMBH) using capillaries treated using the standard method. The labeled protein LssaCAsp- was added to serially diluted NP (GST-NP, GST was used as a negative control) or OsTLPsp- (MBP-fusion protein, MBP was used as a negative control). The initial concentration of each protein was 40 µM with 0.1% (v/v) Tween 20. The KD Fit function of NanoTemper Analysis software version 1.2.214.1 was used for curve fitting and to calculate the dissociation constant (Kd).

β-1,3-glucanase activity assays

β-1,3-glucanase hydrolyzes laminarin by cleaving β-1,3-glucoside bond, thereby generating sugar termini. Enzyme activity was quantified by measuring the rate of reducing sugars production. One unit (U) of β-1,3-glucanase activity was defined as the amount of enzyme that hydrolyses laminarin to generate 1 μg reducing sugars per minute.

The β-1,3-glucanase activity of OsTLP was measured using a beta-1,3-glucanase microplate assay kit (Cat. ASK1028, Bioworld, Bloomington, MN, USA) according to the manufacturer’s instructions. To determine the enzymatic activity of recombinant OsTLP proteins, MBP-fused OsTLPsp- protein and MBP control protein were adjusted to 0.5 mg/mL. The MBP-OsTLPsp- fusion protein was equilibrated in buffers of varying pH for the assay. To determine the enzymatic activity of OsTLP expressed in planta, 0.1 g of rice leaf tissue from either OsTLP over-expressing or wild-type plants was homogenized with 1 mL of assay buffer on ice. The homogenate was centrifuged at 12,000g at 4°C for 10 min, and the resulting supernatant was transferred into a new tube and kept on ice until analysis.

For the assay, 50 μL of the protein solution was mixed with laminarin substrate and incubated at 37°C for 30 min. The reaction was terminated by incubating the reaction tube in a boiling water bath for 10 min. The reaction supernatant was transferred into a microplate and incubated at 90°C for 10 min before measuring the absorbance at 540 nm.

Generation of transgenic plants

The full-length OsTLP gene was PCR-amplified with the primer pair TLP-F/TLP-R (Table S2) and cloned into the plant expression vector pCAMBIA1300 to produce the plasmid pCAMBIA1300-OsTLP. This construct was transformed into seedlings of the Japonica rice cultivar Nipponbare through Agrobacterium-mediated transformation, following the protocol described by Hiei et al. (1994). The ostlp knockout line was generated using CRISPR-Cas9 technology according to established protocols. Thirty independent T0 transformants were obtained and screened by germinating seeds from T2 lines on medium containing 50 μg/ml hygromycin. Hygromycin-resistant homozygous lines were selected for further experiments. Knockout events were confirmed by sequencing with specific primers (Fig S6C). Seeds from homozygous lines transformed with the vector pCAMBIA1300-eGFP were used as negative control.

Statistical analyses

Graphs were generated, and statistical analyses were conducted using Prism 9.0 software (GraphPad Software, San Diego, CA, USA). Data are presented as mean ± standard deviation (SD) from two or three independent experiments. Statistical significance between two groups was assessed using an unpaired Student’s t-test, while comparisons among three or more groups were performed using one-way analysis of variance (ANOVA).

eLife Assessment

This study presents a useful set of experiments to explore how a salivary protein might facilitate planthopper-transmitted rice stripe virus infection by interfering with callose deposition and if fully validated, these findings would significantly advance our understanding of tripartite virus-vector-plant interactions and could be of broad interest to plant science research. The authors provide additional data supporting protein-protein interactions and clarify the transient presence of LssaCA in plants. However, the mechanistic framework remains incomplete, particularly regarding the temporal dynamics of callose function and the sustained effect of LssaCA after virus inoculation. Evidence for the tripartite interaction's functional relevance is still limited, and several critical phenotypic and biochemical details require further substantiation.

Reviewer #1 (Public review):

In this study, the authors identified an insect salivary protein LssaCA participating viral initial infection in plant host. LssaCA directly bond to RSV nucleocapsid protein and then interacted with a rice OsTLP that possessed endo-β-1,3-glucanase activity to enhance OsTLP enzymatic activity and degrade callose caused by insects feeding. The manuscript suffers from fundamental logical issues, making its central narrative highly unconvincing.

(1) These results suggested that LssaCA promoted RSV infection through a mechanism occurring not in insects or during early stages of viral entry in plants, but in planta after viral inoculation. As we all know that callose deposition affects the feeding of piercing-sucking insects and viral entry, this is contradictory to the results in Fig. S4 and Fig 2. It is difficult to understand callose functioned in virus reproduction in 3 days post virus inoculation. And authors also avoided to explain this mechanism.

(2) Missing significant data. For example, the phenotypes of the transgenic plants, the RSV titers in the transgenic plants (OsTLP OE, ostlp). The staining of callose deposition were also hard to convince. The evidence about RSV NP-LssaCA-OsTLP tripartite interaction to enhance OsTLP enzymatic activity is not enough.

(3) Figure 4a, there was the LssaCA signal in the fourth lane of pull-down data. Did MBP also bind LsssCA? The characterization of pull-down methods was rough a little bit. The method of GST pull-down and MBP pull-down should be characterized more in more detail.

Author response:

The following is the authors’ response to the previous reviews

Reviewer #1 (Public review):

In this study, the authors identified an insect salivary protein LssaCA participating viral initial infection in plant host. LssaCA directly bond to RSV nucleocapsid protein and then interacted with a rice OsTLP that possessed endo-β-1,3-glucanase activity to enhance OsTLP enzymatic activity and degrade callose caused by insects feeding. The manuscript suffers from fundamental logical issues, making its central narrative highly unconvincing.

(1) These results suggested that LssaCA promoted RSV infection through a mechanism occurring not in insects or during early stages of viral entry in plants, but in planta after viral inoculation. As we all know that callose deposition affects the feeding of piercing-sucking insects and viral entry, this is contradictory to the results in Fig. S4 and Fig. 2. It is difficult to understand callose functioned in virus reproduction in 3 days post virus inoculation. And authors also avoided to explain this mechanism.

We appreciate your insightful comment and acknowledge that our initial description may not have been sufficiently clear.

(1) Based on the EPG results, we found that LssaCA deficiency did not significantly affect total feeding time, time to first non-phloem phase, or time to first phloem feeding (Fig. S8A-D in the revised manuscript). However, the continuity of sap ingestion was disturbed—the N4 waveform of dsLssaCA SBPHs was occasionally interrupted for brief periods (newly added Fig. S8E in the revised manuscript), likely due to phloem blockage. In the revised manuscript, we have added this analysis to the Result section (Lines 285-291 and 578-587) and provided the EPG procedure in Material and Methods section (Lines 670-680).

(2) We assessed RSV titers immediately post-feeding to confirm the inoculation viral loads (Fig. 2G) and at 3 dpf (Fig. 2H-I) to assess the in-planta effects following viral inoculation. This did not mean that callose functions in virus reproduction at 3 days post viral inoculation. Rather, callose deposition typically occurs immediately in response to insect feeding and virus inoculation. When measuring callose deposition, we allowed insects to feed for 24 h and quantified the callose levels immediately post feeding. The EPG results showed that sap ingestion continuity was disrupted—the N4 waveform of dsLssaCA-treated SBPHs was occasionally interrupted for brief periods (newly added Fig. S8E in the revised manuscript), likely due to phloem blockage. We have reorganized the description to avoid confusion. Please see Lines 139-144 and Fig. S8E for detail.

(1) Missing significant data. For example, the phenotypes of the transgenic plants, the RSV titers in the transgenic plants (OsTLP OE, ostlp). The staining of callose deposition were also hard to convince. The evidence about RSV NP-LssaCA-OsTLP tripartite interaction to enhance OsTLP enzymatic activity is not enough.

We thank the reviewer for this insightful comment.

(1) We constructed OsTLP overexpression and mutant transgenic plants (OsTLP OE and ostlp) and assessed their phenotypes regarding RSV infection levels. Compared with wild-type plants, OsTLP OE plants exhibited accelerated growth, while ostlp plants showed growth inhibition. Following feeding by viruliferous L. striatellus, OsTLP OE plants had significantly higher RSV titers compared with wild-type plants, whereas ostlp mutant plants exhibited significantly lower RSV titers (Lines 221-228 and new Fig. 3I). These results indicate that OsTLP facilitates RSV infection in planta.

(2) The images showing callose deposition staining are representative of 15 images from 3 independent insect treatments. In addition to the staining images, we quantified fluorescence intensity and measured callose concentration by ELISA.

(2) Figure 4a, there was the LssaCA signal in the fourth lane of pull-down data. Did MBP also bind LsssCA? The characterization of pull-down methods was rough a little bit. The method of GST pull-down and MBP pull-down should be characterized more in more detail.

We thank the reviewer for this helpful comment. MBP did not bind LssaCA. We have repeated the pull-down experiment and provide clearer figure with improved results. We have also revised and provided more detailed descriptions of the GST pull-down and MBP pull-down methods. Please refer to Lines 744-774 and Figure 4A for details.

Version 3

Abstract

Plant viruses typically rely on insect vectors for transmission between plants, with insect salivary proteins playing critical roles in this process. In this study, we demonstrate how Laodelphax striatellus salivary carbonic anhydrase (LssaCA) promotes rice stripe virus (RSV) infection in plants. We discovered that LssaCA directly binds to RSV nucleocapsid protein (NP) in insect salivary glands. This LssaCA-NP complex interacts with a rice thaumatin-like protein (OsTLP) that possess endo-β-1,3-glucanase activity potentially degrading callose. Upon binding, the LssaCA-NP complex significantly enhances OsTLP enzymatic activity. We further clarify that both L. striatellus feeding and RSV infection induce callose deposition. The tripartite LssaCA-NP-OsTLP interaction enhances callose degradation, thereby facilitating RSV infection via its insect vector. This study provides new insights into complex virus-insect-plant tripartite interactions mediated by insect salivary proteins, with broad implications for numerous plant viruses transmitted by insect vectors.

Introduction

During plant virus transmission by piercing-sucking insects, most viruses are inoculated into the plant phloem via the insect’s secreted saliva (Arca & Ribeiro, 2018; Conway et al., 2016; Wu et al., 2022). Thus, insect saliva acts as an interface for the virus-insect-host tripartite interaction and can directly promote viral transmission to, and infection of, the host plants (Sun et al., 2020; Wu et al., 2022). However, despite the importance of insect salivary proteins in this tripartite interaction, there is still much to learn about how these proteins enable successful viral infection.

Previous studies have revealed that there are two ways in which insect saliva facilitates viral infection. One is an indirect approach whereby the saliva modulates the host microenvironment at the feeding site, and saliva effectors work together to allow the arthropod to go unnoticed while it feeds on the host plant (Acevedo et al., 2019; Arca & Ribeiro, 2018; Sun et al., 2020). For example, our work indicated that a Laodelphax striatellus mucin protein, LssaMP, enables the formation of the salivary sheath and facilitates the transmission of rice stripe virus (RSV) into the rice phloem (Huo et al., 2022). The study on leafhoppers revealed that the expression of a saliva calcium-binding protein is inhibited by rice gall dwarf virus (RGDV), thus causing an increase of cytosolic Ca2+ levels in rice and triggering callose deposition and H2O2 production. This increases the frequency of insect probing, thereby enhancing viral horizontal transmission into the rice phloem (Wu et al., 2022). The other mechanism by insect saliva to facilitate virus infection is direct regulation, whereby saliva proteins promote virus transmission through specific molecular interactions (Wen et al., 2019). Direct saliva protein-pathogen interactions have been reported in animal pathogens. For example, during transmission of Borrelia burgdorferi by Ixodes scapularis, the saliva protein Salp15 of I. scapularis binds to the bacterial outer surface protein C, which prevents the bacterium from being recognized by the animal immune system. In this way, the saliva protein enables the pathogen to infect the animal host (Ramamoorthi et al., 2005; Schuijt et al., 2008). Although most plant viruses are heavily dependent on insect vectors for plant-to-plant transmission (Gray, 2008), the direct function of insect saliva proteins in mediating virus transmission remains largely uninvestigated.

During sap-feeding, arthropods produce two distinct types of saliva at different stages of the feeding process: gel saliva and watery saliva (Bonaventure, 2012; Lou et al., 2019). The former forms a salivary sheath to provide a smooth path for the stylet penetration (Lou et al., 2019). The latter is mainly secreted into the phloem sieve elements to prevent them from plugging up and suppress plant defense responses. Some salivary components may act as herbivore-associated molecular patterns that can trigger pattern-triggered immunity, and certain salivary effectors may be recognized by plant resistance proteins to induce effector-triggered immunity, etc. (Huang et al., 2017; Ji et al., 2017; Yi et al., 2021).

For phloem-feeding insects, callose deposited on phloem sieve plate and plasmodesmata of sieve elements functions as a defense mechanism by reducing insect feeding and preventing viral movement (Hao et al., 2008; Hipper et al., 2013; Will & Vilcinskas, 2015; Zavaliev et al., 2011; Yue et al., 2022). Callose is a β-(1,3)-D-glucan polysaccharide that is synthesized by callose synthases and degraded by β-(1,3)-glucanases. Plants defend themselves by depositing callose at the sieve plates and plasmodesmata in response to virus infection, whereas viruses counter this defense by activating β-(1,3)-glucanases to degrade callose (Bucher et al., 2001; Hao et al., 2008; Wu et al., 2022; Zavaliev et al., 2011).

RSV is the causative agent of rice stripe disease, a serious disease of rice crops that has occurred repeatedly in China, Japan, and Korea (Xu et al., 2021). RSV is completely dependent on insect vectors for transmission among its host plants, and L. striatellus is the main vector (Xu et al., 2021; Zhao et al., 2018). L. striatellus transmits RSV in a persistent-propagative manner. The virus initially infects the midgut, then disperses from the hemolymph into the salivary glands, and is inoculated into the plant host during L. striatellus feeding (Huo et al., 2022). L. striatellus belongs to the order Hemiptera, whose members mainly feed from sieve tubes through their mouthparts (stylets) that penetrate plant tissues and reach sieve tubes to ingest the phloem sap (Tjallingii, 2006; Van Bel & Will, 2016). RSV is mainly secreted into the rice phloem via the watery saliva (Huo et al., 2022; Wang & Blanc, 2021).

In this study, we identified a molecular interaction among RSV, a L. striatellus saliva protein, and a plant β-1,3-glucanase. The insect saliva protein directly binds to the RSV nucleocapsid protein (NP) and then binds to a rice thaumatin-like protein to activate its β-1,3-glucanase activity. The activation of β-1,3-glucanase helps RSV infection by inhibiting callose deposition in response to viral infection.

Results

L. striatellus salivary carbonic anhydrase directly binds to RSV NP

Insect-secreted watery saliva is inoculated into plant phloem sieve tubes along with RSV (Huo et al., 2022). We identified the salivary proteins that directly interact with RSV to promote viral plant infection. Among the saliva proteins identified in our previous study from RSV-free L. striatellus (Huo et al., 2022), Sap6 (GenBank accession No. RZF48846.1) was found to co-localize with RSV in infected salivary glands, suggesting a potential molecular interaction between them (Fig. 1A and S1).

An L. striatellus salivary carbonic anhydrase (LssaCA) directly binds to RSV NP in salivary glands

A. Immunofluorescence showing co-localization of RSV (red) and LssaCA (green) in L. striatellus salivary glands. Images represent three independent experiments with 30 SBPHs analyzed. Scale bar: 20 μm. psg, principal salivary gland; asg, accessory salivary gland. B. Schematic of LssaCA protein sequence. SP, signal peptide; Carb_anhydrase domain, conserved eukaryotic-type carbonic anhydrase sequence. Triangles indicate seven predicted catalytically active residues. C. Esterase activity of recombinant LssaCA protein expressed in sf9 insect cell and E. coli system. In LssaCA mutant protein, all seven predicted catalytically active residues were replaced. Mean and SD were calculated from 2 independent enzymatic assays. ****, P<0.0001. D. GST pull-down assays demonstrating interaction between LssaCA (GST-LssaCA) and RSV NP (NP-His). GST served as negative control. Anti-GST and anti-His antibodies detected corresponding proteins. E. MST assay showing binding between LssaCA (LssaCA-His) and RSV NP (GST-NP). GST served as negative control. Error bars represent SD. F. Measurement of LssaCA tissure-specific expression by RT-qPCR and western blotting. Third-instar RSV-free nymphs were used to measure the gene expression in gut, salivary gland, fatbody, and remaining tissues (R-body) after dissection of the aforementioned tissues. Adult females and adult males were used to measure gene expression in ovaries and testes, respectively. Mean and SD were calculated from three independent experiments with 5 tissue samples each experiment. LssaCA protein was detected using anti-LssaCA antibodies. G. Western blot analysis of LssaCA protein in RSV-free SBPH salivary glands. dsLssaCA treatment significantly reduces target protein levels compared to controls. LssaCA protein was detected using anti-LssaCA antibodies. H. Western blot analysis showing LssaCA distribution in watery and gel saliva of RSV-free insects. LssaCA protein was detected using anti-LssaCA antibodies. Total saliva proteins were visualized by silver staining. I. Western blot detection of LssaCA proteins in rice phloem immediately following L. striatellus feeding. J. Western blot detection of LssaCA proteins in rice phloem at 1- and 3-days post-feeding (dpf). (I-J) LssaCA protein was detected using anti-LssaCA antibodies. Rubisco large subunit (RBCL) served as protein loading control. K. Analyses of LssaCA levels regulated by RSV infection. Third-instar RSV-free and RSV-infected nymphs were used to measure gene expression (RT-qPCR) and protein accumulation (Western blotting). Mean and SD were calculated from three independent experiments with totally 14 tissue samples each experiment. ns, not significant. All western blot images are representative of three independent experiments.

Sequence homology analysis revealed Sap6 as a carbonic anhydrase-like protein, with a signal peptide located at the N-terminus (amino acids 1-24) (Fig. 1B and S2A). We recombinantly expressed this protein in E.coli or in sf9 cells as a secreted protein. The purified proteins were measured enzymatic activity in hydrolyzing p-nitroethyl acetate. The observed hydrolyzing activity confirmed Sap6 as a carbonic anhydrase (abbreviated as CA, Fig. 1C). Sequence analysis indicated seven residues constituting the enzymatic active center: H111, N139, H141, H143, E153, H166, and T253. We then constructed a mutant with substations of all seven residues (Fig. S2B. H111D, N139H, H141D, H143D, E153H, H166D, and T253E). The mutant protein lost enzymatic activity, indicating the necessity of the enzymatic active center (Fig. 1C). Sap6 was subsequently named L. striatellus salivary carbonic anhydrase (LssaCA).

We performed pull-down and Microscale thermophoresis (MST) assays to confirm the molecular interaction between LssaCA and RSV NP. Pull-down assays revealed a specific interaction between LssaCA and NP (Fig. 1D). MST assays determined that the KD of the interaction between LssaCA and NP was 2.7 ± 2.2 μM, further confirming the molecular interaction (Fig. 1E and S3).

We then performed quantitative reverse-transcriptional PCR (RT-qPCR) to determine the gene expression and western blotting to assess protein distribution. RT-qPCR revealed specific expression of LssaCA in L. striatellus salivary glands (Fig. 1F), while western blotting confirmed the protein’s presence in this tissue (Fig. 1F).

When LssaCA expression was knocked down using LssaCA-specific double-stranded RNA (dsLssaCA) microinjection, LssaCA protein levels in salivary glands decreased dramatically (Fig. 1G). To investigate LssaCA secretion, we allowed RSV-free insects to feed on an artificial diet, collecting watery saliva proteins from the diet solution and gel saliva proteins from the feeding parafilm. Western blotting demonstrated that the LssaCA protein was secreted as a component of watery saliva (Fig. 1H). Furthermore, by measuring LssaCA protein levels in plants immediately after feeding, we confirmed that LssaCA can be secreted into plants (Fig. 1I). To quantify LssaCA protein levels following insect feeding, we allowed 50 RSV-free insects to feed on a confined plant aera for 2 days. Western blotting revealed that LssaCA persisted for at least 3 days post-feeding (dpf), with protein levels declined rapidly between 1 to 3 dpf (Fig. 1J). We also examined the regulation of LssaCA expression by RSV infection. While RSV infection did not significantly alter LssaCA mRNA levels in the salivary glands, the protein levels were significantly elevated in RSV-infected salivary glands (Fig. 1K).

In conclusion, these results indicate that when RSV NP is secreted by L. striatellus, a salivary-gland-specifically expressed protein, LssaCA, can directly bind to RSV and be secreted via saliva.

LssaCA enhances RSV infection in plants

To determine the role of LssaCA in RSV infection of rice plants, 36.8 nL dsLssaCA at 1 ng/nL was microinjected into the hemocoel of RSV-infected 3rd-instar nymphs to interfere with gene expression. At 2 days post-microinjection (dpi), LssaCA mRNA levels were measured by RT-qPCR and compared with the dsGFP control. Results showed that dsLssaCA treatment reduced LssaCA mRNA levels by 80% (Fig. 2A). The two groups of insects were then allowed to feed on healthy rice seedlings (5 insects per rice seedling) for 2 days. RSV NP RNA levels (indicative of RSV titers) in the plants were measured at 14 dpf, when inoculated viruses had infected the plant host, replicated in plant cells, and viral titers were high enough to be detected by RT-qPCR and western blotting. Compared to rice seedlings fed by dsGFP-treated insects, those fed by dsLssaCA-treated insects had significantly reduced RSV titers (Fig. 2B-C), indicating that LssaCA plays an essential role in mediating RSV infection of rice plants.

LssaCA enhances RSV infection in rice plants

A. Gene silencing efficiency determined by RT-qPCR. B-C. Analysis of RSV infection levels in rice seedlings. Each plant seedling was fed upon by 5 infected insects for 2 days, followed by 14 days of cultivation. RSV levels were determined by RT-qPCR (B) and western blotting (C). RSV NP detected by anti-NP polyclonal antibodies indicated viral titers. Rubisco large subunit (RBCL) served as protein loading control. D. Gene silencing efficiency determined by RT-qPCR. E. RT-qPCR analysis of RSV titers in salivary glands. Each dot represents salivary glands from 5 RSV-infected third-instar nymphs. F. RSV titers in L. striatellus saliva determined by RT-qPCR. Each dot represents one saliva sample from 10 insects fed with artificial diet. G. RT-qPCR analysis of RSV titers in rice. Plants were fed upon by infected insects for 24 h (Left: schematic of L. striatellus feeding on specific rice stem sites), with titers assayed immediately post-feeding (Right). Each dot represents an individual rice seedling.. H-I. RSV infection in plants at 3 dpf. Five RSV-infected third-instar L. striatellus nymphs were fed upon healthy rice plants for 2 days, and viral titers in whole plant shoots were measured at 3 dpf by RT-qPCR (I) and western blotting (J). Each dot represents an individual plant. RSV NP detected by anti-NP polyclonal antibodies indicated viral titers. RBCL protein served as a protein loading control. Mean and SD were calculated from 3 independent microinjection experiments. ns, not significant; *, P<0.05; **, P<0.01; ****, P<0.0001. All western blot images are representative of three independent experiments.

To determine at which step LssaCA affected RSV infection, we analyzed the influence of LssaCA-deficiency on RSV titers during several steps of the RSV transmission and infection process, including virus load in salivary glands and saliva, and initial inoculation levels in planta. First, at 2 days after dsLssaCA treatment of viruliferous insects, when LssaCA was significantly downregulated (Fig. 2D), RT-qPCR was used to detect RSV titer in insect salivary glands. Results showed that LssaCA deficiency did not reduce RSV titers in this tissue (Fig. 2E). Second, LssaCA-deficient insects were allowed to feed on an artificial diet, and their saliva was collected for RSV level analysis. Results revealed that LssaCA-deficiency did not reduce RSV titer in the secreted saliva (Fig. 2F). Third, a rice plant was fed with 10 viruliferous insects for 24 h, with insects limited to a specific area, and RSV titers were measured immediately after feeding. We found that LssaCA-deficiency did not affect RSV initial inoculation levels in planta (Fig. 2G). Fourth, we measured RSV infection levels at 3 dpf to assess the in-planta effects following viral inoculation. LssaCA expression was knocked down using dsLssaCA in RSV-infected 3rd-instar nymphs, and 5 insects were allowed to feed on healthy rice plants for 2 days. We then measured RSV titers in whole plant shoot at 3 dpf. Significantly lower RSV titers were observed in the dsLssaCA-treatment group at 3 dpf (Fig. 2H-I).

Collectively, these results suggest that LssaCA promotes RSV infection through a mechanism occurring not in insects or during early stages of viral entry in plants, but in planta after viral inoculation.

LssaCA interacts with rice thaumatin-like protein to increase its endo-β-1,3-glucanase activity

To investigate the mechanism by which LssaCA facilitates RSV infection in plants, we performed yeast two-hybrid screening to identify rice proteins that interact with LssaCA. A rice thaumatin-like protein (XP_025879846.1), designated as Oryza sativa thaumatin-like protein (OsTLP) (Fig. 3A-B and S4, Table S1), was identified. The open reading frame of OsTLP encodes a polypeptide of 327 amino acids with a thaumatin family (THM) domain from residues 27 to 267. A BLAST algorithm-based analysis revealed sequence identities to TLP 1b. The N-terminal 24 amino acids constitute a secretion signal peptide, suggesting OsTLP is a secreted protein. Pull-down and MST assays were performed to confirm the molecular interaction between OsTLP and LssaCA. Pull-down assays showed that OsTLP and LssaCA have a specific interaction (Fig. 3C). The MST assay revealed a KD of 1.3 ± 2.2 μM for the OsTLP-LssaCA interaction, further confirming the molecular interaction (Fig. 3D).

LssaCA interacts with rice thaumatin-like protein (OsTLP) to enhance its endo-β-1,3-glucanase activity

A. Schematic diagram of OsTLP protein sequence. SP, signal peptide; THM, conserved thaumatin family protein sequence. Glyco_hydro_64 domain, glycoside hydrolases of family 64. B. Yeast two-hybrid assay showing interaction between OsTLP and LssaCA. SD, synthetically defined medium; Leu, Leucine; Trp, Tryptophan; His, Histidine; 3’AT, 3-amino-1,2,4-triazole; AD, transcription activation domain; BD, DNA-binding domain. C. Pull-down assays showing interactions between LssaCA (LssaCA-His) and OsTLP (MBP-OsTLP). MBP served as a negative control. Anti-His and anti-MBP polyclonal antibodies were used for protein detection. Western blot images are representative of three independent experiments. D. MST assay showing molecular interactions between LssaCA (LssaCA-His) and OsTLP (MBP-OsTLP). MBP served as a negative control. Error bars represent SD. E. Endo-β-1,3-glucanase activity of purified recombinant OsTLP protein expressed as MBP-fusion protein. MBP was used for normalization. F. Endo-β-1,3-glucanase activity in OsTLP-overexpressing transgenic plants (OsTLP OE) or wild-type plants (WT). Plant total soluble proteins were used for enzymatic activity assays. Units mg-1 means units per mg of plant total soluble proteins. G. Regulation of OsTLP enzymatic activity by LssaCA binding. H. OsTLP enzymatic activity measured at pH 5.0, 5.5, 6.0, 6.5, 7.0 and 7.5 to assess pH dependence. I. RT-qPCR analysis of RSV infection in OsTLP OE, ostlp mutant, and WT plants. RSV titers (NP copy number) were measured at 14 dpf. ns, not significant; *, P<0.05; **, P<0.01; ***, P<0.001; ****, P<0.0001.

Previous studies have shown that TLP orthologs in various plants, including apples, cherries, tomatoes, and barley, possess endo-β-1,3-glucanase activity that degrades callose (de Jesus-Pires et al., 2020). To verify the endo-β-1,3-glucanase activity of OsTLP, we conducted enzymatic assays using recombinantly expressed OsTLP protein. The purified protein exhibited an average activity of 60 units/mg (Fig. 3E). To examine OsTLP activity in planta, we constructed transgenic plants overexpressing OsTLP (OsTLP OE) by using the Japonica rice cultivar Nipponbare (Fig. S5A-B). These transgenic plants showed significantly increased β-1,3-glucanase activity compared to wild-type controls (Fig. 3F).

To determine whether the interaction between LssaCA and OsTLP affects enzymatic activity, we pre-incubated OsTLP with either LssaCA or BSA (control) for 2 h before measuring activity. Pre-incubation with LssaCA significantly enhanced OsTLP activity compared to the BSA control (Fig. 3G). Since CA family proteins catalyze the conversion of CO2 and H2O to HCO₃⁻ and H⁺, potentially lowering pH, we examined the pH dependence of OsTLP activity. Activity assays conducted across a pH range of 5.0-7.5 revealed that OsTLP activity increased at lower pH values (Fig. 3H). However, pH monitoring during OsTLP-LssaCA interaction experiments showed no change from the initial pH of 7.4, indicating that the enhanced enzymatic activity was not due to pH changes but rather to the direct interaction between OsTLP and LssaCA. Collectively, these results suggest that LssaCA may promote RSV infection by enhancing OsTLP’s callose-degrading activity.

OsTLP facilitates RSV infection in planta

To determine the role of OsTLP in RSV infection of rice plants, we constructed ostlp mutant plants (Fig. S5C). Viruliferous L. striatellus were allowed to feed on OsTLP OE plants, ostlp mutant plants or wild-type plants for 2 days, and RSV titers were measured at 14 dpf. RT-qPCR analyses indicated that compared with wild-type plants, plants overexpressing OsTLP had significantly higher RSV titers, while ostlp mutant plants exhibited significantly lower RSV titers (Fig. 3I). These results indicate that OsTLP facilitates RSV infection in planta.

RSV NP-LssaCA-OsTLP tripartite interaction further enhances OsTLP enzymatic activity

Given that LssaCA interacts with both RSV NP and OsTLP, we investigated the possibility of a tripartite interaction. Through pull-down assays, we discovered that when LssaCA was pre-incubated with NP to form an NP-LssaCA complex, this complex retained the ability to bind to OsTLP (Fig. 4A). This result demonstrates a tripartite interaction among these three proteins.

RSV NP-LssaCA-OsTLP tripartite interaction further enhances OsTLP enzymatic activity

A. Pull-down assays showing interactions between OsTLP (MBP-OsTLP) and LssaCA-NP complex. LssaCA was expressed with a His tag, and RSV NP was expressed as a GST-fusion protein. LssaCA and NP were pre-incubated before co-incubation with OsTLP. Anti-His, anti-GST, and anti-MBP antibodies were used to detect corresponding proteins. B. Endo-β-1,3-glucanase activity of OsTLP after co-incubation with LssaCA (LssaCA + GST) and LssaCA-NP complex (LssaCA + NP). LssaCA was pre-incubated with either GST or NP before addition to OsTLP. ****, P<0.0001.

We then explored how the tripartite interaction affects OsTLP enzymatic activity. OsTLP was pre-incubated with either the LssaCA-NP complex or LssaCA alone as a control. After 2 h of incubation, enzyme activity was measured. The results showed that pre-incubation with the LssaCA-NP complex significantly increased OsTLP activity compared to the LssaCA control (Fig. 4B). This indicates that stronger activation of β-1,3-glucanase activity is achieved through the NP-LssaCA-OsTLP tripartite interaction.

LssaCA inhibits insect-feeding and RSV-infection induced callose deposition in rice plants

Callose deposition on sieve plates and plasmodesmata serves as a critical defense mechanism in plants, reducing insect feeding, viral spread, and infection severity (Li et al., 2012; Zavaliev et al., 2011). Given that the LssaCA or LssaCA-NP complex enhances OsTLP’s β-1,3-glucanase activity and potentially promotes callose degradation, we examined callose regulation during feeding by both RSV-free and RSV-infected L. striatellus. To assess feeding-induced callose deposition, we exposed rice seedlings to either RSV-free or RSV-infected L. striatellus for 24 h and analyzed callose levels around the feeding sites. ELISA assays showed that while feeding by RSV-free insects increased callose concentration, feeding by RSV-infected insects triggered an even greater increase (Fig. 5A). Consistent with these findings, microscopic analysis revealed enhanced callose deposition in phloem cells of plants fed upon by RSV-infected L. striatellus compared to those fed by RSV-free insects (Fig. 5B-C and S6). To determine whether this response was localized, we measured callose levels in tissues distant from the feeding sites. No significant differences in callose deposition were observed between control plants and the non-feeding areas (Fig. S7A), indicating that callose accumulation was confined to feeding sites and their immediate vicinity.

LssaCA inhibits L. striatellus-transmitted RSV-induced callose deposition in rice plants

A. Callose concentration quantified by ELISA in leaves of rice plants fed on by RSV-free or RSV-infected L. striatellus. Non-fed plants served as the control (CK). B. Bright blue fluorescence of cross-sections showing callose deposition at feeding sites. Samples were prepared from the leaf phloem of plants that were not fed or were fed on by RSV-free or RSV-infected L. striatellus. Thin sections were stained with 0.1% aniline blue at 24 h post-feeding and examined under a fluorescence microscope. xy, xylem; ph, phloem. Scale bars: 10 μm. C. Average fluorescence intensity of callose deposition measured using ImageJ. Eight to ten random sites per sample were selected for intensity quantification. D–H. Transcript levels of callose synthase genes (gls3, gls4, gls5, gls10) and OsTLP determined by RT-qPCR. Insects were allowed to feed on rice plants for 24 h prior to RNA extraction. I. Callose concentration quantified by ELISA in leaves of rice plants fed on by dsGFP- or dsLssaCA-treated RSV-infected L. striatellus. J. Bright blue fluorescence of cross-sections showing callose deposition at feeding sites. Samples prepared from the leaf phloem of plants fed on by dsGFP- or dsLssaCA-treated RSV-infected L. striatellus. Scale bars: 20 μm. K. Average fluorescence intensity of callose deposition measured using ImageJ. Eight to ten random sites per sample were selected for intensity quantification. Mean and SD were calculated from three independent experiments. ns, not significant; *, p < 0.05; **, p < 0.01; ***, p < 0.001; ****, p < 0.0001.

To evaluate the molecular mechanisms underlying the induced callose deposition, we analyzed transcript levels of callose synthase genes using RT-qPCR. In agreement with the callose deposition patterns, feeding by RSV-free insects evaluated the expression of callose synthase genes, including gsl3, gsl4, gsl5, and gsl10, while RSV-infection further enhanced their upregulation (Fig. 5D–G and S7B-E). These results further support that both insect feeding and RSV infection promote callose deposition. Notably, the transcriptional levels of these genes remained unchanged in non-feeding areas (Fig. S7B-E). We also measured the OsTLP gene expression and found that neither SBPH feeding nor RSV infection significantly induced OsTLP expression (Fig. 5H).

To investigate whether LssaCA influences callose deposition induced by insect-transmitted RSV, we downregulated LssaCA in RSV-infected L. striatellus using dsRNA microinjection. At 2 dpi, when LssaCA expression was significantly reduced, we allowed the insects to feed on healthy plants for 24 h before measuring callose deposition. Our results showed that plants fed upon by LssaCA-deficient insects accumulated more callose compared to those fed upon by control insects (dsGFP treatment) (Fig. 5I-K). These results indicate that LssaCA inhibits callose deposition.

Using Electrical penetration graph (EPG) analysis, we monitored the influence of LssaCA deficiency on the insect feeding behaviors. EPG waveforms can be classified into three different periods: nonpenetration (NP), penetration (N1, N2, N3) and phloem feeding (N4: N4-a, N4-b) periods. The NP waveform indicates a lack of feeding, the N waveform indicates penetration of mesophyll cells to seek the phloem and the N4 waveform indicates digestion of the sap from the phloem. Results indicated that LssaCA deficiency did not affect the time to first non-phloem feeding, total feeding time in the phloem, and the time to first phloem feeding within the 24-h-detection period (Fig. S8A-D). However, the continuity of sap ingestion was disturbed—the N4 waveform of dsLssaCA SBPHs was occasionally interrupted for brief periods (Fig. S8E). Phloem blockage might be a reason for this feeding disruption.

In conclusion, these results demonstrate that both L. striatellus feeding and RSV infection induce callose deposition, which can be suppressed by LssaCA.

LssaCA mediates in planta callose degradation by enhancing OsTLP enzymatic activity

Ostlp mutant plants were used to confirm whether OsTLP degrades the deposited callose induced by insect-transmitted RSV in planta. Both ELISA assays and microscopic analysis showed that compared to wild-type plants, ostlp plants exhibited significantly higher callose levels (Fig. 6A-B).

LssaCA mediates in planta callose degradation by enhancing OsTLP enzymatic activity

A. Callose concentration quantified by ELISA in leaves of wild-type (WT), OsTLP overexpression (OsTLP OE), and OsTLP mutant (ostlp) plants. B. Bright blue fluorescence of cross-sections showing callose deposition in leaf phloem. Plants were not fed on by insects. Scale bars: 10 μm. Average fluorescence intensity of callose deposition was measured using ImageJ. Eight to ten random sites per sample were selected for intensity quantification. xy, xylem; ph, phloem. C. Callose concentration quantified by ELISA in leaves of ostlp plants fed on by dsGFP- or dsLssaCA-treated RSV-infected L. striatellus. D. Bright blue fluorescence of cross-sections showing callose deposition at feeding sites. Samples were prepared from the leaf phloem of ostlp plants fed on by dsGFP- or dsLssaCA-treated RSV-infected L. striatellus. Scale bars: 10 μm. Average fluorescence intensity was quantified using ImageJ. Eight to ten random sites per sample were selected for intensity quantification. Mean and SD were calculated from three independent experiments. ns, not significant; *, p < 0.05. ****, p < 0.0001.

Given that LssaCA mediates callose degradation by enhancing OsTLP enzymatic activity, we hypothesized that when OsTLP is knocked out in plant, LssaCA would not affect callose levels induced by RSV infection. To test this, we allowed dsLssaCA-treated viruliferous L. striatellus or dsGFP-treated control insects to feed on ostlp seedlings. ELISA assays indicated that LssaCA deficiency did not significantly change callose accumulation (Fig. 6C). Microscopic analysis confirmed higher callose deposition in phloem cells of ostlp seedlings, which remained unaffected by LssaCA deficiency (Fig. 6D). These results indicate that OsTLP is required in planta for LssaCA-mediated effects on callose deposition.

Discussion

Insects are known to be vectors of many plant viruses, and some of these plant viruses rely entirely on insect vectors for their plant-to-plant transmission. Insect salivary proteins have been reported to facilitate viral infection of plants through various mechanisms, such as breaking down plant cell walls, inhibiting plant immune defenses and callose deposition, directly binding to viruses to prevent their detection by plant immune systems, and more. In this study, we found that RSV infection of rice plants triggers the deposition of callose. However, this process is reversed by a tripartite interaction involving the central component of an insect salivary protein LssaCA, which subsequently binds to RSV in insect salivary glands and then OsTLP in the plants. The activity of OsTLP’s β-1,3-glucanase is significantly increased when this tripartite interaction occurs; leading to the degradation of callose and thus allowing for efficient RSV infection (Fig. 7).

Proposed model of RSV NP-LssaCA-OsTLP tripartite interaction facilitating callose degradation and promoting RSV infection.

In L. striatellus salivary glands (SG), RSV (red star) binds to LssaCA proteins (blue triangle) via the nucleocapsid protein (NP). Upon delivery into the rice phloem, RSV infection triggers callose synthesis, leading to callose deposition that restricts viral spread. However, LssaCA binds to OsTLP (yellow scissors), enhancing its β-1,3-glucanase activity and promoting callose degradation. The RSV NP-LssaCA complex formed in the insect’s SG can still interact with OsTLP in planta, resulting in significantly higher OsTLP β-1,3-glucanase activity. This enhanced enzymatic activity efficiently degrades callose, thereby facilitating RSV infection.

Callose is a complex polysaccharide, and its accumulation in plant phloem plays an important role in defending against insect attacks and viral spread (Zavaliev et al., 2011). When attacked by piercing insects or viruses, plants activate callose deposition on the sieve plates to occlude the flow of phloem sap, thus discouraging feeding and reducing viral spread. This accumulation of callose in plant phloem can be switched on or off through the action of callose synthases and hydrolases (Li et al., 2012; Shangguan et al., 2018). Studies have shown that expression levels of these glucanase enzymes differ between resistant and susceptible plant species. This suggests that some plants may be more successful at resisting infection due to high levels of glucanase enzymes in their tissues, which allow them to more quickly and effectively form callose barriers against insects and viruses (Hao et al., 2008). Some insects have evolved β-1,3-glucanases enzymes that can degrade the deposited callose, thus enabling them to bypass the plant’s defense mechanisms and continue feeding (Bucher et al., 2001; Hao et al., 2008). In this study, we demonstrate a novel mechanism by which an insect saliva protein binds to a plant β-1,3-glucanases protein, activating the enzymatic activity and promoting callose degradation.

In this study, we found that plants fed on by viruliferous insects exhibited higher callose deposition than those fed on by uninfected insects (Fig. 5A–C), suggesting that viral infection via insect saliva may constitute a dual stress to the plant, inducing enhanced callose deposition that must be overcome for successful insect feeding and viral spread. To investigate the underlying mechanism, we examined the tripartite interaction between LssaCA-NP-OsTLP. We found that this interaction induced significantly higher β-1,3-glucanase activity than the LssaCA-OsTLP interaction alone, suggesting that SBPH and RSV have evolved mechanisms to enhance β-1,3-glucanase activity in order to bypass the callose-mediated defense deposition and facilitate successful viral spread.

Carbonic anhydrase catalyzes the reversible hydration of carbon dioxide according to the reaction CO₂ + H₂O ⇌ HCO₃⁻ + H⁺. This enzymatic activity is essential for maintaining pH homeostasis in biological systems by regulating the balance between CO₂, bicarbonate, and protons. The importance of carbonic anhydrases in pH regulation is exemplified in mosquito larvae, which utilize an extremely alkaline digestive strategy (pH 10.5-11) that is relatively rare in nature, with carbonic anhydrases serving as the central enzymatic system for generating and maintaining bicarbonate/carbonate buffering in the midgut lumen (Linser et al., 2009). Similarly, carbonic anhydrases in arthropod saliva can significantly impact host-pathogen interactions through pH modulation. A recent study elucidated how aphid salivary carbonic anhydrase II (CA-II) facilitates virus transmission, with the enzyme showing significantly higher expression levels in winged morphs compared to wingless counterparts. When secreted into the plant apoplastic space, CA-II catalyzes H⁺ accumulation, resulting in apoplastic acidification from 5.5 to 5.0. This pH reduction enhances polygalacturonase activity, thereby promoting the degradation of demethylesterified homogalacturonan components. Consequently, plants respond by accelerating vesicle trafficking to transport pectin for cell wall reinforcement, a process that simultaneously facilitates viral translocation from the endomembrane system to the apoplastic compartment (Guo et al., 2023). In contrast to these pH-dependent mechanisms, our study demonstrated that LssaCA’s biological function in mediating RSV infection is, if not completely, at least partially independent of its enzymatic activity. We found that although OsTLP enzymatic activity is highly pH-dependent and exhibits increased enzymatic activity as pH decreases from 7.5 to 5.0 (Fig. 3H), the LssaCA-OsTLP interaction occurring at pH 7.4 significantly enhanced OsTLP enzymatic activity without any change in buffer pH (Fig. 3G). These results demonstrate the crucial importance of LssaCA-OsTLP protein interactions, rather than enzymatic activity alone, in mediating RSV infection. While the exact role of LssaCA enzymatic activity in aiding RSV infection remains unclear, the local low-pH plant environment would enhance OsTLP enzymatic activity, suggesting potential synergistic effects between enzymatic and non-enzymatic functions.

In conclusion, this study elucidated a tripartite interaction among a plant virus, an insect saliva protein, and a host phloem β-1,3-glucanase. This interaction inhibited the plant defense of callose deposition and enabled viral phloem transmission. This study provides new insights into the tripartite virus-insect vector-plant interaction, which is relevant to many agriculturally important plant arboviruses whose transmission is facilitated via insect saliva proteins.

Materials and methods

Viruses, insects, host plants, and antibodies

The RSV-free and RSV-infected L. striatellus individuals used in this study were originally captured in Jiangsu Province, China, and were maintained in our laboratory. All plants used to rear L. striatellus were grown in a growth incubator (14.5 cm height, 3.2 cm radius) at 25℃ under a 16-h light/8-h dark photoperiod. Insects were transferred to fresh seedlings every 15 days to ensure adequate nutrition and insect activity.

RSV ribonucleoproteins (RNPs) were purified from RSV-infected rice plants. These purified RNPs were subsequently used as the antigen to immunize rabbits for the production of an RSV-specific antibody. The LssaCA polyclonal antibody was produced using recombinantly expressed LssaCA protein (with a His-tag at the C-terminus) as the antigen. Anti-His, anti-GST, and anti-MBP antibodies were purchased from ABclonal.

Tissue collection

Insect dissection and tissue collection were performed using previously described procedures (Huo et al., 2018). In brief, insects were anesthetized at 4°C for 10 min, and the forelegs were severed at the coxa-trochanter joint using forceps. The hemolymph was expelled and drawn to the tip of clean forceps. The insects were then dissected from the abdomen in pre-chilled PBS buffer. The fat body was collected with forceps, aided by the tension of the liquid. Tissues, including the guts, salivary glands, and ovaries, were washed three times in PBS to remove contaminants from the hemolymph.

Bioinformatic analysis

The signal peptide and transmembrane helix were predicted using tools at the SignalP 5.0 server (https://services.healthtech.dtu.dk/service.php?SignalP-5.0) and the TMHMM-2.0 server (https://services.healthtech.dtu.dk/service.php?TMHMM-2.0). Homology searches of the protein sequence of LssaCA and OsTLP were conducted using the BLAST algorithm at the National Center for Biotechnology Information (http://blast.ncbi.nlm.nih.gov/Blast.cgi). Conserved domains were predicted with SMART (http://smart.embl-heidelberg.de/).

Collection of watery and gel saliva

Saliva was collected as described previously (Huo et al., 2022). Watery saliva was obtained by centrifuging the collected artificial diet at 10,000 ×g for 10 min to remove the solid gel. To collect gel saliva proteins, the membrane was washed twice with PBS. Sheathes adhering to the parafilm were denatured and solubilized in a solution consisting of 8 M urea, 4% v/v CHAPS, 0.1% w/v sodium dodecyl sulfate (SDS), and 2% v/v DTT. The parafilm was incubated on an orbital shaker at room temperature for 45 min. Then, the denatured gel saliva was concentrated to a final volume of 100 µL. Saliva samples were collected from 2,000 insects for silver staining and western blotting analyses.

Measurement of carbonic anhydrase activity

DNA fragments encoding full-length LssaCA and a LssaCA mutant (H111D, N139H, H141D, H143D, E153H, H166D, and T253E) were amplified by PCR. The resulting products were purified and cloned into the vector pET-30a(+) and pFastBac1 vectors to generate His-tagged recombinant proteins. The pET-30a(+)-LssaCA and pET-30a(+)-LssaCA(mutant) plasmids were transformed into E. coli for protein expression. In parallel, the pFastBac1-LssaCA and the pFastBac1-LssaCA(mutant) plasmids were transformed into E. coli DH10Bac for transposition into the bacmid. Recombinant bacmids were then transfected into Spodoptera frugiperda (Sf9) cells using Cellfectin II Reagent (Invitrogen, USA). Recombinant proteins expressed in both systems were purified using Ni Sepharose resin (GE Healthcare, Uppsala, Sweden).

Carbonic anhydrase activity was assessed by measuring esterase activity using p-nitroethyl acetate (p-NPA) as the substrate. LssaCA and its mutant were expressed in E. coli and Sf9 cells with C-terminal His-tag, and the purified proteins were diluted in PBS buffer (pH 7.0) to a final concentration of 1 mg/mL. Reaction conditions followed the method described by Verpoorte et al. (1967), with minor modifications. Briefly, 1 mL of freshly prepared 3 mM p-NPA (dissolved in acetone) was added to 1.9 mL of 100 mM PBS (pH 7.0), followed by 0.1 mL of LssaCA enzyme solution (0.5 mg/mL in 100 mM PBS). Absorbance at 384 nm was measured at the start of the reaction and after 3 minutes. One unit (U) of enzyme activity was defined as the amount of enzyme required to hydrolyze 1 μmol of p-NPA per minute at pH 7.0 and 25 °C. A blank control consisted of 2 mL of 100 mM PBS (pH 7.0) mixed with 1 mL of 3 mM p-NPA.

Immunofluorescence assays

Immunofluorescence assays (IFA) were performed to determine the localization of RSV and LssaCA in L. striatellus salivary glands. The dissected salivary gland samples were fixed in 4% v/v paraformaldehyde at 4°C overnight and then permeabilized in 2% v/v Triton X-100 for 2 h. The samples were incubated with rabbit anti-RSV serum and mouse anti-LssaCA serum (1:250 dilution in PBST/FBS: PBS containing 2% v/v Triton X-100 and 2% fetal bovine serum) for 1 h at room temperature. After washing twice with PBST, the samples were incubated with 1:200 diluted Alexa Fluor 488-conjugated goat anti-mouse antibody and Alexa Fluor 594-conjugated goat anti-rabbit antibody (Invitrogen, USA) for 1 h. After washing three times with PBST, the samples were examined under a Leica TCS SP8 microscope (Leica, Wetzlar, Germany) with the following settings: 1) Alexa Fluor 488 channel: excitation at 488 nm using an argon laser; emission detected with a 493-553 nm bandpass filter. Alexa Fluor 594 channel: excitation at 594 nm using a HeNe laser; excitation detected with a 600-650 nm bandpass filter.

Electrical penetration graph (EPG)

EPG was performed according to previously reported (Huo et al., 2022). L. striatellus individuals were pre-starved for 4 h and then anesthetized for 10 min at 4°C. A gold wire (20 μm diameter, 4 cm long, connected to the EPG probe via a copper nail) was attached to the dorsal thorax of the insect using water-soluble silver glue. When the glue had dried, the probe was connected to the amplifier and the insect was placed on the leaf of a 2-week-old rice plant in a pot. The other amplifier was connected to the soil in the pot with another brass wire. The electronic signals were amplifier, and the digitized data were received by a computer using Style software (Wageningen University, Wageningen, Netherlands). Data were recorded for each insect continuously for 12 h.

Evaluation of callose deposition in rice plants

Five insects were allowed to feed for 24 h on a 2 × 2 cm leaf area of 10-d-old rice plants. After removal of the insects, leaf tissues were collected from both the feeding site and an adjacent undamaged area located 1 cm away (non-fed site). Samples were immersed in 5 mL alcoholic lactophenol solution (1 volume of phenol:glycerol:lactic acid:water in a 1:1:1:1 ratio, mixed with 2 volumes of ethanol), evacuated with a syringe for 15 min, and then incubated at 65℃ for 30 min until the chlorophyll was completely cleared. Cleared leaf samples were rinsed with 50% v/v ethanol, then with water, and then stained in 150 mM K2HPO4 (pH 9.5) containing 0.01% w/v aniline blue for 30 min in the dark. Stained leaf samples were sectioned and mounted in 50% v/v glycerol on slides. The slides were examined under a Zeiss observer Z1 immunofluorescence microscope (Carl Zeiss MicroImaging GmbH, Gottingen, Germany) with excitation at 405 nm. Fluorescence intensity was quantified using ImageJ software. For ImageJ analysis, images were converted to 8-bit grayscale and adjusted for brightness. Phloem-specific regions of interest (ROIs) were outlined using the freehand selection tools. A standardized fluorescence threshold (set at >2× the mean background intensity) was applied to isolate callose signals. Integrated density and area values were used to calculate normalized fluorescence intensity. Identical exposure times were maintained across all independent experiments.

Callose concentrations were quantified using a plant callose ELISA kit (Catalog No. JL46407; Jonln, Shanghai, China), following the manufacturer’s instruction. Leaf tissues from feeding and non-fed sites were homogenized in PBS buffer (pH 7.4), and the homogenate was centrifuged at 500–1000 x g for 20 min. The supernatant was collected, and absorbance at 450 nm was measured. Callose concentrations were calculated using a standard curve.

RNA interference

We conducted RNAi analyses to determine the function of LssaCA in mediating RSV infection and regulating callose deposition. The LssaCA-specific gene fragment was PCR-amplified with the primer pair LssaCA-T7F/LssaCA-T7R (Table S1). Then, dsRNA was synthesized using a commercial kit (T7 RiboMAX Express RNAi System, Promega, Madison, WI, USA) and purified by phenol: chloroform extraction and isopropanol precipitation. Finally, 36.8 nL dsRNA at 1 ng/nL was delivered into the insect hemocoel for gene silencing. Microinjection was performed using a glass needle through a Nanoliter 2000 (World Precision Instruments, Sarasota, FL, USA). The negative control, GFP dsRNA, was synthesized and microinjected following the same protocol.

RT-qPCR

Gene transcript levels were determined by RT-qPCR. The RNA was extracted from L. striatellus tissues using a RNeasy Mini Kit (Qiagen, Hilden, Germany), and from plant leaves using TRIzol. Reverse transcriptional PCR (iScript cDNA Synthesis Kit, Bio-Rad, Hercules, CA, USA) and SYBR-Green-based qPCR (SYBR Green Supermix, Bio-Rad) was performed according to the manufacturer’s protocols. The primer pair used to amplify LssaCA was LssaCA-q-F/LssaCA-q-R (Table S1). Viral RNA copies were measured using the primers NP-q-F/NP-q-R (Table S1). L. striatellus translation elongation factor 2 (EF2) and rice actin were amplified as internal controls to ensure equal loading of cDNA isolated from different samples. The primer pairs EF2-q-F/EF2-q-R and actin-q-F/actin-q-R (Table S1) were used to amplify EF2 and actin, respectively. Water was used as a negative control.

Yeast two-hybrid assays

Yeast two-hybrid screening was performed to identify rice proteins interacting with LssaCA. A high-complexity rice cDNA library was fused to Gal4 AD and transformed into the yeast strain Y187. The LssaCA gene lacking the signal-peptide sequence (LssaCAsp-) was fused to Gal4 DNA-BD and transformed into the yeast strain Y2HGold. The two yeast strains were co-cultured overnight to allow mating to create diploids. After library screening, positive clones were selected on quadruple dropout medium: SD/–Ade/–His/–Leu/–Trp supplemented with X-a-Gal and aureobasidin A (QDO/X/A), and prey plasmids were isolated for sequencing. To further confirm the interaction between LssaCA and OsTLP, LssaCAsp-was cloned into the prey vector pDB-Leu, and the OsTLP gene lacking the signal-peptide sequence (OsTLPsp-) was cloned into the bait vector pPC86. The two plasmids were co-transformed into the yeast stain Mav203 and the transformants were selected on SD/–His–Leu–Trp + 20 mM 3-AT medium.

Pull-down assays

Pull-down assays were performed to determine the molecular interactions between LssaCA and RSV NP, LssaCA and OsTLP, and OsTLP and the LssaCA-NP complex. The LssaCA sequence lacking the signal peptide (LssaCAsp-) was cloned into the pGEX-3x and pET-30a(+) vectors to generate GST-tagged (GST-LssaCA) and C-terminal His-tagged (LssaCA-His) fusion proteins, respectively. Similarly, the full-length NP gene was cloned into the same vectors to produce GST-NP and NP-His. The OsTLP sequence without its signal peptide (OsTLP sp-) was cloned into the pMAL-p2X vector to express an N-terminal MBP-fusion protein (MBP-OsTLP). All recombinant proteins were expressed in and purified from E. coli strain BL21 for subsequent assays.

The interaction between LssaCA and RSV NP was determined using GST pull-down analyses as described previously (Huo et al., 2014). Briefly, GST was used as a negative control. GST-LssaCA or GST protein was bound to Glutathione Sepharose beads (Cytiva, Marlborough, MA, USA) for 2 h at 4 °C. Subsequently, purified NP-His protein was added to the beads and incubated for another 2 h at 4 °C. After centrifugation and five washes, the bead-bound proteins were separated by SDS-PAGE and detected by western blotting with antibodies against GST (Tiangen, Beijing, China) or His-tag (Tiangen, Beijing, China).

Next, MBP pull-down analyses were performed to determine the interaction involving OsTLP, LssaCA, and RSV NP. MBP was used as a negative control. The purified MBP-OsTLP or MBP protein was bound to amylose resin (New England Biolabs, Beverley, MA, USA) for 2 h at 4 °C, and then the LssaCA-His protein was added to the beads. To determine the interaction between OsTLP and the LssaCA-NP complex, a separate reaction was performed in which LssaCA-His and GST-NP proteins were pre-incubated together for 2 h at 4 °C to form the complex before addition to the resin. An equivalent amount of LssaCA-His was used in both reaction systems. After centrifugation and five washes, the bead-bound proteins were separated by SDS-PAGE and detected by western blotting with the MBP antibody (Cell Signaling Technology, Danvers, MA, USA), His-tag antibody (Tiangen, Beijing, China) and GST antibody (Tiangen, Beijing, China).

Microscale thermophoresis (MST) assays

We conducted MST assays to detect the interactions between LssaCA and NP, and between LssaCA and OsTLP. First, 10 μM purified LssaCAsp- protein (His-fusion protein) was labelled with a Monolith NT Protein Labeling Kit RED-NHS (Nano Temper Technologies GMBH, München, Germany) using red fluorescent dye NT-647 N-hydroxysuccinimide (amine-reactive), according to the manufacturer’s instructions. The binding assays were performed on a Monolith NT.115 Microscale Thermophoresis instrument (Nano Temper Technologies GMBH) using capillaries treated using the standard method. The labeled protein LssaCAsp- was added to serially diluted NP (GST-NP, GST was used as a negative control) or OsTLPsp-(MBP-fusion protein, MBP was used as a negative control). The initial concentration of each protein was 40 µM with 0.1% (v/v) Tween 20. The KD Fit function of NanoTemper Analysis software version 1.2.214.1 was used for curve fitting and to calculate the dissociation constant (Kd).

β-1,3-glucanase activity assays

β-1,3-glucanase hydrolyzes laminarin by cleaving β-1,3-glucoside bond, thereby generating sugar termini. Enzyme activity was quantified by measuring the rate of reducing sugars production. One unit (U) of β-1,3-glucanase activity was defined as the amount of enzyme that hydrolyses laminarin to generate 1 μg reducing sugars per minute.

The β-1,3-glucanase activity of OsTLP was measured using a beta-1,3-glucanase microplate assay kit (Cat. ASK1028, Bioworld, Bloomington, MN, USA) according to the manufacturer’s instructions. To determine the enzymatic activity of recombinant OsTLP proteins, MBP-fused OsTLPsp- protein and MBP control protein were adjusted to 0.5 mg/mL. The MBP-OsTLPsp- fusion protein was equilibrated in buffers of varying pH for the assay. To determine the enzymatic activity of OsTLP expressed in planta, 0.1 g of rice leaf tissue from either OsTLP over-expressing or wild-type plants was homogenized with 1 mL of assay buffer on ice. The homogenate was centrifuged at 12,000g at 4°C for 10 min, and the resulting supernatant was transferred into a new tube and kept on ice until analysis.

For the assay, 50 μL of the protein solution was mixed with laminarin substrate and incubated at 37°C for 30 min. The reaction was terminated by incubating the reaction tube in a boiling water bath for 10 min. The reaction supernatant was transferred into a microplate and incubated at 90°C for 10 min before measuring the absorbance at 540 nm.

Generation of transgenic plants

The full-length OsTLP gene was PCR-amplified with the primer pair TLP-F/TLP-R (Table S2) and cloned into the plant expression vector pCAMBIA1300 to produce the plasmid pCAMBIA1300-OsTLP. This construct was transformed into seedlings of the Japonica rice cultivar Nipponbare through Agrobacterium-mediated transformation, following the protocol described by Hiei et al. (1994). The ostlp knockout line was generated using CRISPR-Cas9 technology according to established protocols. Thirty independent T0 transformants were obtained and screened by germinating seeds from T2 lines on medium containing 50 μg/ml hygromycin. Hygromycin-resistant homozygous lines were selected for further experiments. Knockout events were confirmed by sequencing with specific primers (Fig. S5C). Seeds from homozygous lines transformed with the vector pCAMBIA1300-eGFP were used as negative control.

Statistical analyses

Graphs were generated, and statistical analyses were conducted using Prism 9.0 software (GraphPad Software, San Diego, CA, USA). Data are presented as mean ± standard deviation (SD) from two or three independent experiments. Statistical significance between two groups was assessed using an unpaired Student’s t-test, while comparisons among three or more groups were performed using one-way analysis of variance (ANOVA).